Brain dynamics and spatiotemporal trajectories during threat processing

Version 1

Abstract

Abstract

In the past decades, functional MRI research has investigated task processing in largely static fashion based on evoked responses during blocked and event-related designs. Despite some progress in naturalistic designs, our understanding of threat processing remains largely limited to those obtained with standard paradigms with limited dynamics. In the present paper, we applied Switching Linear Dynamical Systems to uncover the dynamics of threat processing during a continuous threat-of-shock paradigm. Importantly, unlike studies in systems neuroscience that frequently assume that systems are decoupled from external inputs, we characterized both endogenous and exogenous contributions to dynamics. First, we demonstrated that the SLDS model learned the regularities of the experimental paradigm, such that states and state transitions estimated from fMRI time series data from 85 regions of interest reflected both the proximity of the circles and their direction (approach vs. retreat). After establishing that the model captured key properties of threat-related processing, we characterized the dynamics of the states and their transitions. The results revealed that threat processing benefits from being viewed in terms of dynamic multivariate patterns whose trajectories are a combination of intrinsic and extrinsic factors that jointly determine how the brain temporally evolves during dynamic threat. Finally, we investigated the generalizability of the modeling approach. The successful application of the SLDS model, trained on one paradigm to a separate experiment illustrates the potential of this approach to capture fMRI dynamics that generalize across related but distinct threat-processing tasks. We propose that viewing threat processing through the lens of dynamical systems offers important avenues to uncover properties of the dynamics of threat that are not unveiled with standard experimental designs and analyses.

Introduction

Historically, systems neuroscience has sought to develop experimental paradigms where an effect of interest is varied while as many other variables remain fixed in an attempt to isolate the impact of the former while controlling for the latter. Such time-tested strategy has provided deep insights into brain and behavior. However, with the advent of neurotechnologies allowing the study of animals in more freely-behaving conditions, more naturalistic paradigms have been introduced Hunt et al. (2021); Testard et al. (2024). At the same time, in human neuroimaging, researchers have defended the idea that dynamic and/or naturalistic paradigms are needed to advance our understanding of brain and behavior in a way that is not entirely constrained by rigid paradigms Hasson and Honey (2012); Sonkusare et al. (2019); Finn et al. (2020); Lee Masson and Isik (2021); Grall and Finn (2022). While such ideas resonate with those from ecological psychology and ethological ap-proaches Gothard et al. (2018); Mobbs et al. (2018); Evans et al. (2019), they introduce considerable challenges in the analysis of brain data and associated behaviors. If there are no specific trials or blocks in an experiment, what should constitute the unit of analysis? If the paradigm is more openended and lacks specific conditions (e.g., “attend to the visual stimulus” vs. “attend to the audio stimulus”), how should brain and behavior be investigated?

A promising approach is the use of unsupervised methods that consider time series dynamics Anderson and Fincham (2014); Vidaurre et al. (2016); Baldassano et al. (2017); Linderman et al. (2016); Taghia et al. (2017); Pandarinath et al. (2018); Townsend and Gong (2018); Morioka et al. (2020); Wu et al. (2021); Singh et al. (2022); Song et al. (2023); Xu et al. (2023); Vahidi et al. (2024); Chen et al. (2024). In particular, Hidden Markov Models (HMMs) can be applied to time series data and partition it into “states” in time. For example, in one study, continuous narratives were presented and state transitions were identified from fMRI signals Baldassano et al. (2017). Notably, state boundaries corresponded to human-annotated ones better than chance, and the match between model and human boundaries increased for more “high-level” brain regions (e.g., angular gyrus). Another study attempted to determine mental states while participants solved novel mathematical problems Anderson and Fincham (2014). Model states were interpreted as reflecting “encoding”, “planning”, “solving”, and “responding” stages based on fMRI data. Although the authors found evidence for the detected states, the study also highlights some of the challenges of applying unsupervised methods to complex paradigms. The estimated duration of the final responding state correlated well with the measured time participants spent entering their responses (r=0.53), providing some validation of that state. However, validation of the other mental states remained challengingAnderson and Fincham (2014).

Overall, despite encouraging results Anderson and Fincham (2014); Vidaurre et al. (2016); Bal-dassano et al.(2017); Taghia et al. (2017, 2018); Meer et al. (2020); Wu et al. (2021); Song et al. (2021, 2023), the extent to which state-space models successfully parse fMRI data into meaningful mental/brain states remains poorly understood. Critically, while state-space models like HMM are able to segment time series data, they do not enable discovery of properties of state dynamics. In particular, how do brain signals evolve during a brain state that is maintained for a period of time? In systems neuroscience progress has been made in this context by studying brain signals via dynamical systems theory Yu et al. (2005); Deco et al. (2008); Shenoy et al. (2013); Remington et al. (2018); Robson and Li (2022). In particular, it has been suggested that attractor dynamics plays important computational functions Miller (2016); Khona and Fiete (2022). However, comparable progress in fMRI research has lagged behind, perhaps because of the slowly evolving nature of the hemodynamic response (but see John et al. (2022)).

Threat processing is typically studied in terms of discrete events and/or blocks that correspond to experimental conditions (but see Mobbs et al. (2007, 2009); Murty et al. (2023); Levitas and James (2024)). For example, a conditioned threat is presented and associated evoked brain responses determined Fullana et al. (2016). But in many real-world situations, conditions involving threat unfold temporally in a less-discrete manner Fanselow and Lester (1987). In the present investigation, we investigated brain dynamics during a continuous threat-of-shock paradigm Meyer et al. (2019); Limbachia et al. (2021) (Fig. 1). We sought to characterize the evolution of brain signals in terms of trajectories in multidimensional space, a framework at times called “computing with trajectories” Buonomano and Maass (2009). Instead of responses to discrete events possibly observed across multiple brain sites, we determined multivariate and distributed patterns of activity with shared dynamics.

Determining shared dynamics during threat processing. Switching linear dynamical systems (SLDS) were used to model fMRI time series data (Yt) from a set of brain regions of interest (ROIs). The framework assumes that time series data can be segmented into a set of discrete states. The model represents brain signals in terms of a set of latent variables (Xt). For each state k the temporal evolution of the system is specified via a linear dynamical system with both intrinsic and input-related components. In the diagram, the system starts in state j (white star) and transitions to state k (the colored patches in the middle represent the subspaces associated with the two states). In state k, the system evolves according to the dynamics matrix Ak and input contributions (Vk). Overall, as states switch temporally, so does the corresponding linear dynamical system governing the system’s trajectory.

We employed a Switching Linear Dynamical Systems (SLDS) framework that estimates a gener-ative model of the data with both states and transitions like HMMs Ackerson and Fu (1970); Barber (2006); Fox et al. (2008); Linderman et al. (2019). In addition, each state is described in terms of a linear dynamical system that specifies the trajectory of the brain during the state. Importantly, unlike studies in systems neuroscience that frequently assume that systems do not receive external inputs (see discussion in Khona and Fiete (2022)), we estimated both endogenous and exogenous components of the dynamics. Our approach demonstrated that we can recover key properties of threat-related processing in the brain Grupe and Nitschke (2013), while establishing novel properties of threat and safety dynamics. More generally, the approach can be applied to a broad range of continuous and/or naturalistic fMRI paradigms investigated in humans.

A final issue we investigated was that of model generalizability: how well does a model estimated in one paradigm generalize to related experiments? Examining this question is of importance because even statistically significant fits do not necessary generalize well to unseen data, not to mention to related but different paradigms. However, if an SLDS model captures meaningful spatiotemporal regularities associated with behavioral/brain states in a given dataset, the SLDS states should provide good fits to unseen data from a second, related experiment. Here, we investigated the extent to which the SLDS model that was fit to the moving circles paradigm generalized to a second, continuous paradigm involving threat and safety conditions.

Results

Continuous threat processing switches between states

Participants watched two circles of different colors on the screen, at times moving close to each other, at times moving apart in a smooth but unpredictable fashion Meyer et al. (2019); Limbachia et al. (2021). Upon circle collision, the participants received mild but unpleasant electrical stimulation together with an aversive sound. To enhance unpredictability and hence anxious apprehension by participants, the movement of the circles included multiple instances of “near misses” in which the circles nearly touched before retreating from each other.

Time series data were analyzed according to the SLDS framework (Fig. 1). Briefly, an SLDS estimates a model of the data using a set of unobserved (also called latent) variables assumed to govern system dynamics. The latent space is typically of considerably smaller dimension relative to the original data space (here 10 latent dimensions and 85 ROI time series, respectively). SLDSs consist of one linear dynamical system model per state, such that the trajectory of the system is governed by a single dynamical systems at at time. Like HMMs, an SLDS segments the data into a fixed number of states, such that the transitions between states follows a Markov process. At a given time, a single state is considered to be “active”. Overall, SLDS models provide a piecewise linear approximation of (potentially nonlinear) dynamics. Time series data were obtained from 85 regions involved in threat-related processing, as defined in a prior study Murty et al. (2022).

The results reported below were obtained with 92 participants. A separate group of 30 participants was used to define hyperparameters, including the number of states (K = 6, see supplementary methods and Fig. S10) and the dimensionality of the latent space (D = 10, see supplementary methods and Fig. S9). Assessing statistical significance when complex models are applied to multiparticipant data requires estimating variability of the estimates across participants. We employed a bootstrap procedure such that an SLDS model was estimated for each bootstrap sample, allowing us to estimate how SLDS parameters varied across the population (see Methods), thus allowing generalization beyond the current sample.

Model states are synchronous with the experimental paradigm

We evaluated if the states estimated by the model reflected experimental stimuli. For example, when the circles were far from each other, participants were “safe”, in contrast to when the circles were rather close and could progress to a collision. We encoded the input based on 20 bins defined in terms of circle proximity and direction: A1-A10 for approach, R1-R10 for retreat (see Methods, top inset in Fig. 2). When the circles approached each other, A1 indicated the farthest distance and A10 indicated that they collided; when they retreated from each other, R10 indicated that the circles collided and R1 indicated the farthest distance. We determined the associations between brain states and stimulus categories (i.e., bins) by calculating the probability of a brain state given the stimulus, P (state | stimulus) (see Methods). The probabilities are shown in Fig. 2A (e.g., P (STATE | 2 A10) = 0.63), where values exceeding those expected by chance are indicated in red outline. A total of five out of six brain states were significantly associated with specific stimuli categories. Examination of Fig. 2A can be used to provide a qualitative description of the estimated brain states. For example, STATE 2 was maximally observed when the circles were approaching and then collided, and thus reflects shock delivery (as well as peri-shock periods). As another example, STATE 5 was observed when the circles were in close proximity but did not touch. This state was particularly interesting given that the experiment was designed such that multiple “near miss” events would occur during the session, so as to enhance participants’ state of apprehension when the circles approached each other.

Brain states, state transitions, and input stimuli. Stimuli were categorized into 20 bins based on circle distance and movement direction: A1-A10 for approach, R1-R10 for retreat (1 indicates circles are farthest and 10 indicates circle collision). (A) Table entries indicate the probability of being in a state given the input stimulus category. (B) Table entries indicate the probability of a state transition given the input stimulus category. In both tables, cells highlighted in red indicate states/transitions significantly associated with the corresponding stimulus category (p < 0.05, corrected for multiple comparisons). The color scale indicates probability.

We found that brain states were observed under specific stimulus conditions. But as the stimulus conditions change, do states transition in a systematic fashion? To evaluate this question, we determined the probability of a state transition given a specific stimulus category (Fig. 2B, e.g., P ([STATE 2 ↦ STATE 1] | A10) = 0.88; see Methods). Fig. 2B shows state transitions that were significantly associated with a stimulus category. For example, STATE 2 (centered on the collision event) transitioned to STATE 1 (“post-shock”) with very high probability for both stimulus categories A10 and R10 (i.e., when the circles collided). In Fig. 2B, to facilitate understanding, we ordered the rows by starting with the STATE 1 to STATE 3 transition (from “post-shock” to “not close”) and then chained the transitions in a manner that reflected the movement of the circles. Thus, the second row shows STATE 3 ↦ STATE 5 (from “not close” to “near miss”), and so on. For ease of reference, we summarize state descriptions in Box 1.

Together, brain states and state transitions were systematically related to external stimuli, demonstrating that the model captured important properties of the experimental paradigm in an unsu-pervised manner.

Model states capture threat processing

We reasoned that if states detected by the model reflect known mental states, they should be associated with brain activity with coherent spatial patterns consistent with those observed with standard fMRI designs (such as block and event-related designs). SLDS states and state transitions are defined in a latent space of dimensionality considerably lower than the original brain data space (here 85 brain regions). Thus, to determine brain maps associated with specific states, we performed linear regression with a regressor that was on during the state in question and off otherwise (see Methods). Fig. 3 illustrates some of the states (see Fig. S12 for all states). For STATE 5, consistent with the “near miss” interpretation, mutiple brain areas linked to increased anxious apprehension were positively engaged, including the anterior insula and cingulate cortex Mobbs et al. (2009, 2010); Alvarez et al. (2011); Grupe and Nitschke (2013); Somerville et al. (2013); Meyer et al. (2019); Hur et al. (2020); Murty et al. (2022, 2023). STATE 4, which tended to follow STATE 5, exhibited a deactivation of these regions, consistent with the notion that participants experienced relative safety following a near collision Schiller et al. (2008); Grupe and Nitschke (2013); Somerville et al. (2013).

Voxelwise state activity maps and state contrast maps. State activity maps for STATE 4 and STATE 5 and the contrast of the two states. Maps were corrected for multiple comparisons by thresholding voxels at p < 0.001 and at the cluster level at p < 0.05.

SLDS state descriptions

STATE 1: “post-shock”. Observed right after the circles collided; only state that followed STATE 2 in Fig. 2B.

STATE 2: “shock/peri-shock”. Observed during shock and peri-shock inputs.

STATE 3: “not near”. Observed when the circles were not near.

STATE 4: “not near”. Observed when the circles were not near and similar to STATE 3 in terms of inputs; STATE 4 followed STATE 5 when, after a near-miss collision, the circles continued to move apart from each other.

STATE 5: “near miss”. Observed when circles approached, came close to colliding, and then retreated.

We further investigated the spatial properties of brain states by contrasting states. For example, the contrast of STATE 5 vs. STATE 4 revealed a pattern of activation very similar to that obtained in the literature for the contrast of higher vs. lower levels of threat. In particular, the contrast identified sectors of ventromedial PFC and posterior cingulate that have been linked to relative safety Schiller et al. (2008); Grupe and Nitschke (2013).

Temporal evolution of activity

In the threat paradigm, as the circles moved continuously, we expected that brain signals would also vary temporally. If SLDS states captured meaningful information, state transitions should be associated with relevant activity change. To evaluate this possibility, we determined signal evolution during temporal windows centered around state transitions (see Methods). Fig. 4 shows signal evolution in the same order as the state transitions of Fig. 2B in two key regions involved in threat-related processing, the dorsal anterior insula and the ventromedial PFC. For example, the transition from STATE 2 to STATE 1 reflected transitioning from a shock event to post-shock period. Accordingly, activity in the dorsal anterior insula, which is very sensitive to threat level Mobbs et al. (2009, 2010); Alvarez et al. (2011); Grupe and Nitschke (2013); Somerville et al. (2013); Meyer et al. (2019); Hur et al. (2020); Murty et al. (2022, 2023), underwent a drastic decrease. Likewise, when STATE 5 transitioned to STATE 4, activity in the dorsal anterior insula decreased, consistent with a transition from higher to lower threat and findings about responses in these two regions. In contrast, the transition from STATE 3 (“not close”) to STATE 2 (peri-shock/shock) was associated with a sharp increase in activity strength in the dorsal anterior insula.

Region and network responses during state transitions. Vertical lines indicate the time of state transition. State transitions are arranged in a chained manner: STATE 1 ↦ STATE 3 ↦ STATE 5 ↦ STATE 4 ↦ STATE 3 ↦ STATE 2 ↦ STATE 1 ↦ STATE 4 ↦ STATE 5 to facilitate continuity in reading responses across state transitions. Error bars correspond to the 95% confidence interval based on the standard error of the mean across participants.

SLDS models states and state transitions in a latent space, which makes it possible to investi-gate activity dynamics (as in Fig. 4) in regions not used for estimating the models, because once a sequence of states is determined one can examine activity in such regions at those times. Furthermore, activity can be determined at different spatial units, including ROI, voxel, or voxel cluster. Here, we probed average signals from two resting-state networks that are engaged during threat-related processing, the salience network which is particularly engaged during higher threat, and the default network which is engaged during conditions of relative safety. As anticipated, signal evolution during the transition STATE 5 ↦ STATE 4 was similar to that observed for the dorsal anterior insula and the ventromedial PFC, although signal evolution for the default network deviated to some extent from that of the ventromedial PFC (note that the ROIs and networks discussed here spatially overlapped; they were not selected to be independent).

Overall, the model captured regularities of the experimental paradigm reflected in systematic changes in activity levels across the brain. For example, state transitions STATE 3 ↦ STATE 5, STATE 4 ↦ STATE 3, and STATE 4 ↦ STATE 5 were associated with approaching circles (increased threat) and increased responses in the dorsal anterior insula and salience network following the state transition. Similarly, the transition STATE 5 ↦ STATE 4 was associated with retreating circles (decreased threat) and increased responses in the ventromedial PFC and default network following the state transition. Furthermore, responses of the right anterior insula and salience network during STATE 2 (“shock/peri-shock”) ramped up to the highest activity level with the STATE 3 ↦ STATE 2 transition from low to very high threat followed by a sharp decay with the STATE 2 ↦ STATE 1 transition that reflected the end of shock delivery.

Threat dynamics

Each state is characterized by a state-specific trajectory in latent space. Such state dynamics are defined by an intrinsic component and an external-input component, which additively define the system’s evolution (see Fig. 1 and Methods). In this section, we characterized the dynamics of threat processing based on state trajectories.

Intrinsic state dynamics evolve towards fixed-point attractors

First, we examined intrinsic state dynamics, namely the component of the dynamics that is separate from external input contributions. Does intrinsic dynamics converge to an attractor or does it diverge away from a fixed point? It is possible to test the stability of dynamics by evaluating properties of the dynamics matrix (matrix A; see Methods). We found that the intrinsic dynamics of all states studied were attractors (see supplementary Fig. S13).

An attractor state corresponds to a point in latent space. Such vector can be visualized as a map, which we call an “attractor spatial map” by projecting it onto the original brain data space in terms of ROI activity on a brain surface (Fig. 5; see Methods). For example, STATE 5 corresponded to “near miss” events and an attractor map with increased responses in the anterior insula/cingulate cortex and decreased responses in the ventromedial PFC and precuneus. The attractor for STATE 4 (“not near”) exhibited an opposite organization consistent with the context of increased safety/lower threat.

Attractor maps. State attractors projected onto the space of regions of interest (85 ROIs) visualized on a brain surface map. At each ROI, the color scale represents activity strength at the attrator’s fixed point. ROI boundaries are marked in black.

Given that the SLDS model learns a generative model of the fMRI data during the experimental paradigm, state trajectories are based on both endogenous and exogenous contributions. How do these trajectories evolve relative to the state attractor? To investigate this question, we determined the vector field defined by the state’s endogenous dynamics, where vectors represent the direction and magnitude of how a system evolves if started at specific points (Fig. 6). Starting at a given point, trajectories evolved for multiple time steps in the overall direction of the attractor but after some time transitioned to other states (the green part of the trajectory indicates the mean lifetime of each state). The plots describe how, in a given state, the effective trajectory was determined by a combination of the endogenous flow and perturbations provided by the input. For example, in STATE 2 (shock/peri-shock) the vector field exhibited less curvature than several other states but trajectories were quickly steered away from the attractor because of inputs.

Evolution of state-specific trajectories. Trajectories were determined in the latent space and projected onto a two-dimensional vector field for illustration (coordinate axes are specific to each state). The average trajectory across participants, starting at the white star, is initially shown in green and switches to red when the majority of trajectories (across participants) switch to another state. Whereas the trajectory includes endogenous and exogenous contributions to temporal evolution, the vector field represents the endogenous contribution only. The gray arrows indicate the effect of the state’s endogenous dynamics matrix, showing the direction and magnitude of evolution after a single time step. The blue cross indicates the state’s fixed point attractor; star indicates the start of the trajectory. PC: principal component.

Exogenous contributions to state trajectories

Following the qualitative description above (Fig. 6), we next quantified the contributions of external inputs in steering system trajectories. For each state, we determined the state’s centroid, namely, the centroid coordintate of all locations visited by the system during a specific state. Such geometrical location provided a reference position to investigate the input contributions. We considered the state centroid as reference location because we sought to determine how the inputs steered trajectories relative to the “most typical” position of a state.

We considered two scenarios. First, when the system was in state i, the input could drive the trajectory towards state i’s centroid or away from it (Fig. 7A). The “effective contribution” was determined by considering both direction and magnitude components (see Methods). We investigated the stimulus categories defined previously (A1—A10 and R10—R1). Consider STATE 1 (post-shock), for example (Fig. 7B). When the stimulus was R10 or R9 those inputs tended to push the system in the direction of STATE 1’s centroid; when the stimulus was R8, the trajectory was pushed away from the centroid by the input but very weakly. In addition, for STATE 2 (shock/peri-shock), we found that stimuli A8 to R10 pushed the system toward the centroid, but stimulis R9 moved the trajectory away from it. Overall, Fig. 7B displays how inputs steered state trajectories in the direction of the centroid (green cells) or away from it (purple cells).

Effect of external inputs on steering state trajectories and driving state transitions. (A) Left: Trajectory at t − 1 in state i. Inputs (colored arrows) could perturb the trajectory (translucent paths of the same color) by steering the evolution towards the state’s centroid (green) or away from it (pink). When the input directed it away from the state’s centroid, the input could push the system to switch into state j (red). Right: The input effect was measured via the cosine of the angle ϕ between the input vector and the line joining Xt−1 and state centroids. (B and C) Input effects with rows and columns representing states/state-transitions and input categories, respectively. Only states with significant association with inputs and only significant state transitions are shown (see Fig. 2). Cells with significant effects are highlighted in blue (p < 0.05, corrected for multiple comparisons).

Next, we investigated how inputs contributed to state transitions by steering trajectories in the direction of a new state’s centroid (see Methods). Consider the STATE 2 ↦ STATE 1 transition (i.e., from shock to post-shock). Input perturbations R10 and R9 provided the strongest push of the trajectory in the direction of STATE 1. In addition, when the system was in STATE 3 (“not near”), multiple inputs tended to direct the system’s evolution in the direction of STATE 2 (“shock/perishock”). Finally, when the system was in STATE 4 (“not close”) and the circles were approaching, inputs pushed the system in the direction of STATE 5 (“near miss”).

Taken together, the results show how external perturbations contributed to the evolution of state trajectories and state transitions.

States explain threat processing in another threat-related experiment

If SLDS models capture spatiotemporal regularities associated with behavioral/brain states, determining states in a paradigm (such as the present moving circles paradigm) that shares important elements with a different experiment should be reflected when fitting the data of the latter. In other words, how well does a model estimated in one paradigm generalize to related experiments?

To test model generalizability, we applied the SLDS model estimated on the moving circles paradigm (call it MOV) to a separate experiment and dataset (call it ESC for “escape”) that investigated threat and safety processing. In the ESC experiment, participants manually controlled a rabbit icon using a joystick in a threat-avoidance task. The participant’s goal was to reach a safety area on the opposite end of the screen while avoiding a pursuing devil icon moving at varying speeds. If caught, participants received a mild but unpleasant electrical stimulus. Once a gate was successfully crossed, the participant entered a safe area (no shock was possible) where they remained for a few seconds (see Fig. 8A). Overall, the paradigm induced anxious apprehension during the chase phase and relief during safety.

Testing generalizability of the SLDS model. (A) Experimental design of the “escape” (ESC) paradigm (data from a separate group of participants). Each trial started with a chase period (pink), where the participant controlled the rabbit icon to move towards the safety area (to their right) while being actively pursued by a threat icon. If the player entered the safety area without getting caught by threat, the safety gate closed, starting a safety period (green). Brain states were analyzed during windows centered around the time of entering the safe region (indicated by the vertical dashed line). (B) Table entries indicate the probability of being in a state given the time step relative to the window. Time steps marked in red and green indicate chase and safety periods respectively. (C) Table entries indicate the probability of a state transition given the time step. In both tables, cells highlighted in red indicate states/state transitions significantly associated with the window’s time step (p < 0.05, corrected for multiple comparisons). The color scale indicates probability values. No hemodynamic shift was applied to the tables in B and C. (D) Average fMRI response in the left Bed nucleus of the Stria Terminalis (BST). Error bars correspond to the 95% confidence interval based on the standard error of the mean across participants.

Model parameters inferred from the MOV dataset were used to decode states in the ESC dataset. We evaluated the extent to which MOV states could explain fMRI time series data of the esc design. To do so, we focused on a ten time-step long temporal window anchored around the event of entering safety (from −6.25 seconds to 5.0 seconds, with t = 0 indicating the time of entering safety; see Methods). We determined the probability of being in a given MOV state given the time step (Fig. 8B). The results indicated that STATE 5 (“near miss” in MOV) was expressed during the chase period; STATE 3 (“not near” in MOV) was detected from t = 0 seconds to the end of the temporal window. We also evaluated whether the system switched from STATE 5 to STATE 3 around t = 0 seconds by determining the probability of observing the transition STATE 5 ↦ STATE 3 as a function of time point (Fig. 8C; see Methods).

In the MOV paradigm, STATE 5 was associated with a state of apprehension likely experienced by participants during near-miss events. This is consistent with the finding that STATE 5 was detected during the chase phase of the ESC paradigm. STATE 3 (“not near” in MOV) occurred during the safety period (in ESC) and the transition from STATE 5 ↦ STATE 3 at t = 0. These findings indicate that the model trained on MOV was able to generalize in meaningful ways to the new paradigm.

To facilitate interpretation, no hemodynamic shifts were used in displaying the results shown in Fig. 8. Our choice was motivated by the fact that the ESC paradigm contained a discrete event at t = 0, namely, the transition into safety. However, one concern is that STATE 3 was already detected at safety onset, whereas safety-related processing triggered at t = 0 should be expressed in fMRI responses with some lag given that any responses triggered at t = 0 should translate to fMRI responses with a lag of around 2 seconds and peak around 4-6 seconds post event Boynton et al. (1996). Further analysis, however, helped explain the observed results, as outlined next. Average fMRI responses during the analyzed time window revealed a consistent response pattern across multiple brain regions sensitive to sustained threat in prior studies in the literatureMobbs et al. (2009); Davis et al. (2010); Murty et al. (2023), as illustrated for the region of interest of the Bed nucleus of the Stria Terminalis (BST; Fig. 8D), a very important brain region involved in sustained threat processing (Davis et al., 2010). As shown, the BST response started decreasing before the end of the chase period, not at the end, suggesting that apprehension levels during the chase period decreased before the end of the chase period. To gain further insight concerning the response pattern of sustained-threat processing regions, we investigated the movement pattern of the chasing devil, which revealed a potential problem with the original experimental design. We programmed the chasing speed of the devil icon to be variable, but consistent across trials. In doing so, there was a consistent icon deacceleration around the halfway point of the chase, which unintentionally made the final phase of the chase less apprehension-inducing. In all, we believe the expression of STATE 3 (“not near” in MOV) right at entry into the safety area, as well as the transition from STATE 5 ↦ STATE 3, are actually consistent with a mental state of relative safety, as also reflected in the activation pattern in the BST (and other brain regions).

Discussion

In the present paper, we applied Switching Linear Dynamical Systems (SLDS) to uncover the dynamics of threat processing during a continuous threat-of-shock paradigm. First, we demonstrated that the SLDS model learned the regularities of the experimental paradigm, such that states and state transitions estimated from fMRI time series data from 85 ROIs reflected both the proximity of the circles and their direction (approach vs. retreat). After establishing that the model captured key properties of threat-related processing, we characterized the dynamics of the states and their transitions. The results revealed that the understanding of threat processing is enhanced by viewing it in terms of dynamic multivariate patterns whose trajectories are a combination of intrinsic and extrinsic factors that jointly determine how the brain temporally evolves during dynamic threat.

Past work has employed state-space models such as HMMs to estimate states from brain signals in an unsupervised fashion. Previous fMRI studies detected states with blocked design Vidaurre et al. (2018); Yamashita et al. (2021), where brain states are expected to align closely to the experimental structure. In continuous experimental paradigms, establishing brain states that meaningfully track experimental variables is considerably more challenging. In fact, in some studies employing HMMs with more unconstrained or naturalistic experiments, states were not explicitly mapped to stimulus events Meer et al. (2020); Song et al. (2021, 2023). A recent interesting study applied HMMs to cell recordings in the rat PFC and found states that correlated with hide- and-seek behaviors, as well as states corresponding to previously undetected behaviors Bagi et al. (2022). In general, although HMM models applied to brain data have shown to be a promising strategy to uncover unlabeled states, further work is needed to assess their performance in continuous paradigms in a manner that rigorously establishes the correspondence between states and external stimuli and/or behavior.

Critically, standard HMMs provide limited information about state dynamics, which is essen-tial in understanding temporally unfolding brain processes supporting complex mental states and behaviors, such as the varying levels of threat and safety studied here. To address this gap, we applied the SLDS framework to our continuous threat-of-shock paradigm. There have been a few applications of SLDS models to brain data Petreska et al. (2011); Linderman et al. (2019); Glaser et al. (2020), but as far as we know only one fMRI study investigating a blocked working memory task Taghia et al. (2018). As one of the challenges with unsupervised state-space methods is that there is no ground truth concerning the states, here we first established that the model estimated states and transitions that captured the structure of the continuous paradigm, namely, circle proximity, shock events, and post-shock period. Of the six states detected by the model, five captured properties of the stimulus in a statistically significant manner. For example, the probability of being in STATE 2 given that the circles collided (input: A10) was 0.63, consistent with a state indicative of a “shock/peri-shock” period. State transitions were also associated with specific stimuli. More generally, the states captured the structure of the paradigm, which consisted of periods of increased and decreased threat given the proximity of the circles and the unpleasant shock delivered upon collision.

Further evidence that the states captured important features of threat-related processing was obtained by determining voxelwise brain maps for both individual states and state contrasts. For example, the contrast of STATE 5 (“near miss”) vs. STATE 4 (“not close”) revealed a pattern of activation very similar to that obtained in the literature for the contrast of higher vs. lower levels of threatMobbs et al. (2009, 2010); Alvarez et al. (2011); Grupe and Nitschke (2013); Somerville et al. (2013); Meyer et al. (2019); Hur et al. (2020); Murty et al. (2022, 2023). In particular, the contrast identified sectors of ventromedial PFC and posterior cingulate that have been linked to relative safety Schiller et al. (2008); Somerville et al. (2013); Grupe and Nitschke (2013). In addition, state transitions reflected notable changes in activity levels, such as the transition of the lower-threat STATE 4 to the higher-threat STATE 5. Taken together, our analyses demonstrated that the model captured key properties of the continuous paradigm, building on previous studies using HMMs. Vidaurre et al. (2018); Meer et al. (2020); Song et al. (2023); Baldassano et al. (2017); Williams et al. (2022).

Our central goal was to discover and characterize the dynamics of continuous threat-related processing. From the perspective developed here brain regions jointly participate in state-specific trajectories. An important property of the SLDS framework is that it estimates state dynamics based on separate and additive endogenous and exogenous contributions. Determining the intrinsic dynamics is particularly informative because it uncovers how a brain state would evolve in the absence of additional inputs. In particular, does the state trajectory tend toward a fixed attractor, or does it evolve in some other more complex manner? Based on the properties of the endogenous state dynamics matrix, we determined that indeed all states behaved as attractors. But what do these attractors represent in terms of brain activity? To visualize the corresponding patterns, we determined “attractor maps” by projecting the model attractor (determined in the latent space) onto the brain space (times series data from ROIs). The attractor maps captured essential properties of the states determined by the model. For example, STATE 1 and STATE 2 maps exhibited inverse activity patterns, consistent with the fact that they largely reflected the post-shock and the shock/peri-shock periods, respectively. In addition, comparison of STATE 5 (“near miss”) and STATE 4 (“not near”) shows how distributed patterns with shared dynamics are involved in mental states of higher and lower anxious apprehension.

The endogenous dynamics can be further understood as providing a flow field that channels the temporal evolution of the system in the absence of inputs. However, in our continuous threat-of-shock paradigm, state trajectories were also determined by the inputs acting at each time step. Thus, visualizing the trajectory in terms of vector fields uncovered both contributions to system behavior. In combination with the information obtained from attractor maps, these results provide novel ways to understand the roles of key brain regions involved in threat-related processing. In addition to the qualitative information provided by vector field trajectories, we quantified the input contributions to states and state transitions. We determined how inputs contributed to trajectories in a state’s subspace and to transitions between states. For example, when the system was in STATE 2 (shock/peri shock), the input perturbation R10 steered the system in the direction of STATE 1 most vigorously, showing that soon after the circles collided, the brain tended to readily transition to a state of decreased activity in the insula and medial PFC. Overall, the analysis revealed how inputs steered a state’s trajectory toward or away from the state’s centroid, as well as how inputs directed a trajectory’s evolution in the direction of a new state.

Work on dynamics of neural circuits in systems neuroscience typically assumes that the target circuit is driven only by endogenous processesKhona and Fiete (2022). In other words, the external inputs do not vary significantly over time and therefore do not perturb state trajectories. While such a scenario may be adequate in some experimental paradigms, the assumption is problematic in cases where inputs vary continuously. In the present study, we did not assume that the brain was isolated from the environment. Instead, by parsing the dynamics in terms of endogenous and exogenous contributions, we were able to not only study intrinsic state attractors but also to characterize external inputs as perturbations that drove trajectories along meaningful directions, such as pulling or pushing inputs toward or away from state centroids. Additionally, we also identified input stimuli that effectively steered evolution in the direction of other state subspaces. We propose that our approach can be profitably employed in studying neuronal circuit dynamics in systems neuroscience more broadly.

We believe the generalizability results presented in this study offer a valuable contribution to the ongoing discussion about the applicability of naturalistic paradigms in neuroscience. As highlighted in the Introduction, there is a growing interest in using more dynamic and ecologically valid experimental designs to advance our understanding of brain function. The successful application of the SLDS model, trained on the moving circles paradigm, to the separate escape experiment illustrates the potential of this approach to capture neural dynamics that generalize across related but distinct threat-processing tasks. This finding addresses a key challenge in naturalistic neuroscience: establishing that models derived from one paradigm can meaningfully explain brain activity in another. By showing that states identified in one threat-related task could be meaningfully mapped onto brain activity during a different threat-avoidance scenario, this study provides initial evidence that the SLDS framework can extract robust, task-general representations of threat processing, rather than idiosyncratic features of a particular experimental paradigm. Such cross-paradigm validity (see also Song et al. (2023)) will be crucial for building a more comprehensive and ecologically relevant understanding of how the brain processes and responds to dynamic threats in real-world scenarios.

Here, we also discuss some challenges and potential shortcomings of our study. The application of state-space models, including slds, to hemodynamic data is challenging because of the low-pass nature of the response. Specifically, the response to a brief event builds up to a peak around 4-6 seconds post onset and decays slowly over the next 4-8 seconds Boynton et al. (1996). In the current study, we assumed a constant hemodynamic lag of approximately 5 seconds. Although our paradigm involved continuous movement of the circles, both shock events and transitions from approach to retreat (and vice versa) were discrete in nature. These events call for careful interpretation of the results given the potential for blending of hemodynamic responses. However, our experiment was designed such that most periods of circle approach and retreat lasted for 6-9 seconds so as to minimize consecutive approach/retreat transitions Limbachia et al. (2021). Thus, the concern about the slow nature of the hemodynamic response was somewhat alleviated in the present context. In general, future studies are needed to develop SLDS approaches that can explicitly consider the properties of the hemodynamic response. Another limitation of the framework developed here is that we assumed, as in all state-model applications, that the underlying systems is well approximated by a series of states. With SLDS models this is less severe an assumption because one can consider the (possibly nonlinear) behavior of a system as modeled by a series of linear dynamical systems that approximate the overall behavior. Nevertheless, it will be valuable to investigate other approaches that address this concern, including considering that multiple states can coexist at a time Greene et al. (2023).

In the present paper, we applied Switching Linear Dynamical Systems to uncover the dynamics of threat processing during a continuous threat-of-shock paradigm. First, we demonstrated that the model learned the regularities of the experimental paradigm, such that states and state transitions estimated from fMRI time series data from 85 ROIs reflected both the proximity of the circles and their direction (approach vs. retreat). After establishing that the model captured key properties of threat-related processing, we characterized the dynamics of the states and their transitions. The results revealed that threat processing can profitably be viewed in terms of dynamic multivariate patterns whose trajectories are a combination of intrinsic and extrinsic factors that jointly determine how the brain temporally evolves during dynamic threat. We propose that viewing threat processing through the lens of dynamical systems offers important avenues to uncover properties of the dynamics of threat that are not uncovered with standard experimental designs and analyses.

Methods

Moving Circles Dataset

Participants

One hundred and twenty-six participants (63 females, ages 18–30 years; average: 20.87, STD: 2.56) with normal or corrected-to-normal vision and no reported neurological or psychiatric disease were recruited from the University of Maryland community. The project was approved by the University of Maryland College Park Institutional Review Board and all participants provided written informed consent before participation. Data from four participants were not used due to technical issues. For the original publication of these data, see Limbachia et al. (2021).

Out of the 122 participants, we used 30 to perform preliminary testing of the framework and defining model hyperparameters, including the number of states and the dimensionality of the latent space(see following subsections and supplementary). In order to avoid circularity in our analyses, data from this exploratory set were not used further. Thus, all results reported here are based on the held-out set of 92 participants.

Procedure and Stimuli

The experiment was designed to study the role of proximity on threat-of-shock processing. Participants watched two circles of different colors on the screen, at times moving close to each other, at times moving apart in a smooth but unpredictable fashion. Upon circle collision, the participants received a mild but unpleasant electrical stimulation together with an aversive sound. The movement of the circles was defined such that there were segments of approach and retreat that lasted between 2 and 9 seconds, with most segments lasting more then 6 seconds. To enhance unpredictability and hence anxious apprehension by participants, the movement of the circles included multiple instances of “near misses” in which the circles nearly touched (distance as close as 1.5 circle diameters between circle centers) before retreating away.

The experiment consisted of six 8-minute runs, each having two 3-minute blocks, during which the circles moved. The participants experienced on average 4 circle collisions and 7 near misses per run. All runs used different motion paths, but the visual stimulus was fixed across participants. Finally, visual stimuli were presented using PsychoPy(http://www.psychopy.org/) and viewed on a projection screen via a mirror mounted to the scanner’s head coil. For further details about the experiment, the reader may refer to Meyer et al. (2019); Limbachia et al. (2021).

Escape Dataset

Participants

A total of 57 subjects (24 females, ages 18-54; average: 20.5, STD: 1.9) were recruited from the University of Maryland, College Park, community. All subjects had normal or corrected-to-normal vision and reported no neurological disease or current use of psychoactive drugs. They provided written informed consent before participating in the study and were paid immediately after the experiment. The study was approved by the University of Maryland, College Park, Institutional Review Board.

Procedure and Stimuli

The experiment sought to study threat processing in presence of “imminent danger” followed by a period of safety. A total of eight 512 second-long runs were employed, each having 13.8±1.6 trials. Each trial began with a chase period that a followed by a safety period if the participant avoided being caught. Participants controlled a rabbit icon (“player”), which they could move on the screen at constant speed using a joystick. The player started at the left hand side of the screen and their objective was to reach a safety area located at the right end of the screen while avoiding a devil icon that pursued the player at time-varying speeds. Participants received an unpleasant but not painful electrical stimulation if the devil caught the player before the player entered the safety region. If the player successfully entered the safety area, this period lasted 6-10 seconds.

Regions of Interest

Time series data used for both datasets consisted of fMRI time series from 85 cortical and subcortical ROIs (defined as the average across voxels for each ROI). The ROIs were defined in a prior study by our group and intended to capture brain regions involved in threat-related processing Murty et al. (2022).

Switching Linear Dynamical Systems

Switching Linear Dynamical Systems (SLDSs) are an example of a latent-variable framework that expresses the dynamics of observed signals using a set of unobserved variables assumed to govern system dynamics. The latent space is typically of considerable smaller dimension relative to the original data space (here 10 and 85, respectively). SLDSs consist of one linear dynamical system model per state, such that the trajectory of the system is governed by a single dynamical system at at each time step. Switching between states is assumed to follow a Markov process. Overall, SLDS models provide a piecewise linear approximation of (potentially nonlinear) dynamics.

Consider activity Yt ∈ ℝN, t = 1, 2, …, T (T time steps is the duration of a typical run) from N = 85 brain ROIs during an experiment described using M input variables Ut ∈ ℝM, t = 1, 2, …, T. The model represents brain signals Yt in a low dimensional subspace as a set of D latent variables Xt ∈ ℝD, t = 1, 2, …, T (D < N). Observed BOLD activity activity is related to the latent variables using a linear transformation, such that where C ∈ ℝN×D is the observation matrix, d ∈ ℝN is a bias term, Σv ∈ ℝN×N is a noise covariance matrix and 𝒩 (µ, Σ) is a multivariate Gaussian distribution with mean µ and covariance Σ. The expression above is called the observation step.

Latent trajectory dynamics is modeled using a set of K linear dynamical systems that are indexed by a time varying switching variable Zt ∈ {1, 2, …, K}, which we call the state. At each time step t only the linear dynamical system corresponding to the state Zt drives the temporal evolution of system, such that where is the linear dynamics matrix, is the input matrix reflecting the experimental paradigm variables, is a bias term and is a noise covariance matrix. The expression above is called the dynamics step.

State transition follows a modification of the Markov property, where the probability of switching to state Zt from the previous state Zt−1 depends on the previous values of the state Zt−1 and the latent variables at t − 1, Xt−1: where rj ∈ ℝD and Q ∈ RK×K is the Markovian transition matrix, whose ith row and jth column element indicates the transition probability from state i to state j. This expression above is called the state switching step. rj is a weight vector (indexed by the state label, j) that specifies the additional dependency of Zt on Xt−1 (on top of the Markov dependence of Zt−1 on Zt), allowing state transitions based on the location of the trajectory in the latent space. SLDS with the above additional dependency is also known as recurrent SLDS Linderman et al. (2016) and is shown to be more suitable to explain realistic timeseries than SLDS without the recurrent dependencies Linderman et al. (2019). Note that, on removing rj from the above equation the state transition probability is the same as that of a Markov model.

A group-level SLDS model was estimated based on the data of 92 participants that participated in the moving circles experiment by using a variational Laplace Expectation Maximization algorithm Linderman et al. (2016). Specifically, all of the following were estimated:. In this manner, we obtained model variables Xt and Zt. We employed the SLDS implementation provided in the ssm python library (https://github.com/lindermanlab/ssm).

Modeling experimental stimuli as SLDS inputs

To encode the experimental paradigm, we categorized the stimuli based on the proximity and direction of motion of the moving circles. The instantaneous proximity was calculated as the normalized Euclidean distance between the circles, ranging from 0 (farthest) to 1 (touching). Direction was calculated based on the discrete difference of proximity: +1 (approach) or -1 (retreat). Proximity values were quantized into 10 equal-sized bins ([0.0, 0.1), [0.1, 0.2), ⋯, [0.9, 1.0]), one per binning per direction, resulting in 20 stimulus categories: A1, A2, ⋯, A10 for approach; R1, R2, ⋯, R10 for retreat). The stimulus conditions were thus encoded as a 20-dimensional binary vector Ut at each time step t, with a 1 corresponding to the active category and 0 elsewhere. The number of stimulus categories was arbitrarily defined, but we verified that key results were robust to the number of input bins (see Supplementary Fig. S14 and S15).

For a state k, the SLDS model represents the contribution of input on dynamics using the term VkUt (Eq. 2), where Ut encodes the stimulus and Vk maps the input into the latent space. Note Vk is state-specific, allowing distinct representations of the same stimulus category per state.

Assessing statistical significance using bootstrapping

Data from the held-out set of 92 participants of the moving circles dataset) were used to report findings of the study and determine statistical significance, which was determined via bootstrapping Efron and Tibshirani (1994). A total of 500 bootstrap samples of the dataset were generated by sampling participants with replacement, and an SLDS model was fit on each bootstrap iteration.

Because state labels are not necessarily preserved across bootstrap samples, we realigned them using a combination of the Hungarian algorithm Kuhn (1955) and k-means clustering. To do so, for each bootstrap sample, state parameters were concatenated to form K = 6 “large vectors” θ, one for each state. The resulting vectors were then clustered into K clusters. The K cluster centroids provided representative vectors θ for each state. In this manner, aligning state labels across bootstrap samples amounted to assigning each bootstrap sample’s θ vectors to the K cluster centroids. The assignment problem was determined by the Hungarian assignment algorithm, such that the bootstrap sample’s state θ vectors were maximally aligned to the K cluster centroids.

To determine the number of bootstrap iterations, we generated samples in steps of 100 until the estimates of the above mentioned θ centroids were stable. Specifically, the elements of each centroid did not change (within 10−5 precision) upon adding bootstrap samples. The procedure was stopped at 500 bootstrap samples.

Calculating the probability of a state given a stimulus

To determine P (state input), we defined a state by input table. Each entry corresponded to the proportion of time a state Zt assumed a specific value z given that the input Ut assumed a specific value u. This counting procedure can be briefly summarize in the (somewhat awkward) expression where 𝟙 (·) represents an indicator function that takes value 1 when the operand is true and 0 otherwise. Given the hemodynamic delay (approximately 2 secs to initiate a response and 4-6 secs to peak; Boynton et al. (1996)), we used a time shift of 3 time steps (3.75 seconds) in considering the relationship between SLDS state and stimulus category. That is, to facilitate interpretation, we assumed that the effect of circle proximity (as encoded in terms of the discrete stimulus categories) would be expressed in the fMRI timeseries 3 time steps later.

We tested the statistical significance of each entry in the state by input table as follows. To account for sampling variability of the sample we studied, we computed a table for each iteration of the bootstrap, allowing us to estimate the variability of our results across the population (group level). Similarly, we obtained surrogate tables, one for each bootstrap resample, by randomly shifting the state sequences by random offsets. In this manner, the relative order of the states was randomized with respect to order of stimuli but the other properties were maintained (transition structure in the sequences). By obtaining bootstrap and surrogate tables; for each table cell corresponding to a given state and stimulus, we obtained a population-level distribution of table values and a null distribution.

Based on these results, each table entry mean was compared to the mean of the correspond-ing null distribution via a paired Wilcoxon signed rank test at p < 0.05. To correct for multiple comparisons across the number of states and stimulus categories (6 × 20), Bonferroni correction was applied.

Calculating the probability of a state transition given a stimulus

State transitions were linked to the stimuli using an analogous procedure as that to map states to stimuli. We created a table with 8 state transitions of interest (see supplementary methods, Fig. S11) by 20 stimulus categories. Each table entry was calculated by counting the number of times a given state transition was observed given a stimulus category. Formally, this counting procedure can be succinctly as follows: for a row r representing transition state i ↦ state j and column u, the corresponding table entry measures the probability where 𝟙(·) represents an indicator function. Once again, this is calculated by considering the hemo-dynamic delay. Finally, statistical inference was performed using a similar approach as described previously for calculating probabilities of states given stimuli.

State spatial maps

We sought to characterize the spatial organization of brain states: for a given state k, what is the typical pattern of activity throughout the brain. This could be accomplished by using multiple regression with variables for each state: 1 when the state is on, 0 otherwise. We performed multiple regression in this manner in a voxel-wise manner. Note that this was possible because once the states are estimated (based on 85 ROIs), the regression can be applied at any spatial level desired (ROI or voxel). The regression model can be expressed mathematically as where 𝟙 (Zt = k) is an indicator function denoting whether state k is active at time t, βk is the corresponding regression coefficient, and εt is an error term. For notational simplicity, terms accounting for confounds like scanner drift and head motion have been omitted from the above expression but were included in the regression model.

State maps were evaluated statistically based on the bootstrap approach discussed previously. First, state activity maps were determined for every participant, separately. A group-level map was then obtained by averaging across participants. This procedure was performed for every bootstrap sample, allowing us to estimate sampling variability. Similarly, a set of null state maps was generated by repeating the same procedure but with randomly permuted state sequences. To identify statistically significant voxels, we performed a paired t-test between the participant-based boot-strapped maps and the null maps. To account for multiple comparisons, we employed a cluster-based approach Friston et al. (1994). Voxels were first thresholded at p < 0.001, and clusters with at least c = 18 voxels were deemed statistically significant at the cluster level. To determine the minimum cluster size c, we employed null state maps. The cluster size c was such that only 5% of the clusters in the null state maps would be detected, corresponding to a 5% false-positive rate.

For contrast maps, an analogous approach was employed, but at the individual level contrast maps were determined by computing βi − βj, where i and j are two states in question.

Responses during state transitions

Responses at a given ROI were evaluated during temporal windows centered around state transitions. For transitions from state i to state j, we chose windows where the first half of the time steps were labelled as state i and the final time steps as state j. We chose windows of length 8 time steps such that they were long enough to reveal activity changes but also observed frequently enough to generate a reliable estimate. For a given state transition, participant-level responses were determined by averaging responses across all qualifying time windows. To estimate variability, group-level responses were determined by averaging participant-level responses for each bootstrap sample.

We also determined responses at the “network” level, which were obtained by averaging ROI responses across all ROIs inside the network. We considered the salience (ventral attention) and default networks, as defined in the Schaefer 100-region 7-network parcellation Schaefer et al. (2017).

State specific intrinsic dynamics and fixed points

Each state’s dynamics in the latent space was modelled using a linear dynamical system, which evolves as follows: To characterize the intrinsic dynamics exclusively, assume that the input to the system is zero across time: Assuming that the dynamics are stable and approach an equilibrium point as a function of time (also known as a “fixed point” or attractor), we obtain where I is a D × D identity matrix. All trajectories in the neighbourhood of the fixed point (also called the basin of attraction) will eventually converge towards the fixed point.

Testing for stable intrinsic dynamics

The dynamics of a linear dynamical system is determined by the eigenvalues of the dynamics matrix Ak. In particular, the dynamics is stable if all eigenvalues lie inside a unit circle on the complex plane (refer to Luenberger (1979); Robinson (2012) for detailed accounts on stability of linear dynamical systems). Therefore it suffices to test whether norm of the largest eigenvalue is less than unity. Bootstrapping was again used to evaluate the statistical significance of the results. We determined largest eigenvalues for all bootstrap samples, and calculated the p-value as the fraction of the eigenvalues with absolute value greater than 1. States with p-value < 0.05 were considered to follow stable dynamics. Bonferroni correction was applied to correct for multiple comparisons across K states.

Visualizing state attractors on the brain

Given an attractor state i, its fixed point is a vector in the D-dimensional latent space. For the purpose of visualizing state attractors on the brain, we projected them onto the original data space (the time series space of 85 ROIs) using the observation step of the SLDS model (Eq. 1). where is state i’s attractor in ROI space. State attractors in ROI space were estimated across bootstrap samples, and the average pattern was visualized. The average attractor pattern was thresholded at each ROI (p < 0.05) based on a Wilcoxon signed-rank test (p < 0.05); that is, by comparing the activity levels to zero. Bonferroni correction was applied to correct for multiple comparisons across K states and 85 ROIs.

State trajectories and vector fields

For each state, trajectories were determined first in the latent space (D = 10). We averaged (across participants) all windows with 10 time steps starting at a specific state k. Note that trajectories correspond to Eq. 4 and consider both endogenous and exogenous terms. At each time step of the average trajectory, a trajectory could continue along a state or switch to a state j ≠ k. The state label for a time step along the trajectory was assigned to the most typical state at that time step; i.e., the modal state. We colored the trajectory green when the state label was k and red otherwise (state j ≠ k). To illustrate the trajectory, we plotted it in 2D. To do so, we fit principal component analysis on the 10 points constituting the average trajectory and projected it onto the plane spanned by the top two principal components.

Trajectories were plotted on vector fields that represent how a particle would flow under the endogenous dynamics (i.e., based on Ak). This allowed us to visualized how external inputs contributed to the evolution of the system (note that, by definition, without inputs trajectories would converged to the state’s attractor). At each point on the 2D plane, the vectors represent the trajectory evolution for a single time step. Vector field points were sampled uniformly on the 2D plane. The points were projected to the latent space using the top two principal components above, evolved according to endogenous dynamics rule (Eq. 5), and projected back onto the 2D plane again using the top two principal components. Finally, the state’s fixed point attractor (a point in the latent space) was also projected onto the 2D plane in the same manner.

Quantifying contribution of inputs in state dynamics

As described above, state k’s trajectory follows the following dynamical rule given in Eq. 4, with the term VkUt representing contribution of external inputs as a vector in latent space. To investigate the contribution of external inputs to the latent trajectory, we determined how inputs steered the system’s trajectory. For each state, we determined the state’s centroid, namely, the vector Xc,k that represented the centroid coordinate of all locations visited by the system during state k. Such geometrical location provided a reference position to investigate the input contributions.

Given a state, the external input at a given time can either drive latent trajectories towards the state centroid or away from it. To quantify input effects, we measured the length of the projection of the input vector along the direction connecting the position of the state trajectory and the state centroid: where ⟨·, ·⟩ indicates the inner product between two vectors and ‖·‖ indicates the Euclidean norm. Thus, positive (negative) leff values indicate that the input pushes state trajectories towards (away from) the state centroid. We focused on inputs associated with the given state as defined throughout the paper (methods for mapping states to stimuli, see Methods). Accordingly, the input effects summarised in a table where the (i, j)th element represents the effect of input j on trajectories in state i.

Tables were constructed as previously such that each bootstrap iteration generated one instance of the table. The results shown correspond to the average value across bootstrap samples. Statistically, we tested whether the average input effects were significantly different from zero by conducting a Wilcoxon’s signed-rank test at each table entry (p < 0.05). Bonferroni correction was applied to correct for multiple comparisons across K = 6 states and 20 input categories.

Quantifying contribution of inputs in state transitions

Next, we quantified the contributions of input stimuli to state transitions. Specifically, for particular state transitions i ↦ j, we determined the extent to which input stimuli contributed to such transitions. As above, leff was used to measure the effect of the external input vector along the direction joining the position of the state trajectory and state j’s centroid (in place of state i’s centroid). Positive values indicate that the input pushed the latent vector from state i towards state j.

Tables were constructed as previously, such that each bootstrap iteration generated one instance of the table. The results shown correspond to the average value across bootstrap samples. Statistically, we tested whether average input effects were significantly different from zero by conducting a Wilcoxon’s signed-rank test at each table entry (p < 0.05). Bonferroni correction was applied to correct for multiple comparisons across 8 state transitions of interest and 20 input categories.

Testing SLDS model generalizability

After fitting the SDLS model based on the data of the moving circles (MOV) paradigm, we investigated if the model states generalized to another threat-related paradigm, namely, the escape (ESC) experiment described above (see Fig 8A). Specifically, after fitting the data to the first dataset, we determined the variables Xt and Zt for the ESC dataset. ESC time series data was analyzed for temporal windows centered around the time step when the player entered safety (rounded to the nearest data sample). Windows straddled between chase and safety periods, from t = −6.25 to t = 5.0 with t = 0 being the cross-into-safety point.

To determine the best-matching MOV states for the ESC time series data, we quantified P (state time step for all combinations of state by time steps see Fig. 8. Each value was computed as the normalized frequency of observing a state Zt at a given time step twin of the window. Formally: where 𝟙 (·) represents the indicator function.

We determined the statistical significance of the probability values as follows. To determine the probabilities that would be observed by chance, we determined surrogate tables (states by time step) by shifting the MOV state sequences by random offsets (analogous to the approach used in mapping states to stimuli in MOV, see Methods), which provided a null distribution of probability values. Each probability value was then compared to the mean of the corresponding null distribution via a paired Wilcoxon signed rank test at p < 0.05. To correct for multiple comparisons across the number of states and time steps (6×10), Bonferroni correction was applied. In a similar fashion, the probability of a state transition from STATE 5 ↦ STATE 3 was tested for statistical significance (analogous to mapping state transitions to stimuli in MOV, see Methods).

eLife Assessment

Using highly sophisticated switching linear dynamical systems (SLDS) analyses applied to functional MRI data, this study provides important insights into network dynamics underlying threat processing. After identifying distinct neural network states associated with varying levels of threat proximity, the paper provides compelling evidence of intrinsically and extrinsically driven contributions to these within-state dynamics and between-state transitions. Although the findings could be made more biologically meaningful, this work will be of interest to a wider functional neuroimaging and systems neuroscience community.

Reviewer #1 (Public review):

Summary:

The manuscript uses state-of-the-art analysis technology to document the spatio-temporal dynamics of brain activity during the processing of threats. The authors offer convincing evidence that complex spatio-temporal aspects of brain dynamics are essential to describe brain operations during threat processing.

Strengths:

Rigorous complex analyses well suited to the data.

Weaknesses:

Lack of a simple take-home message about discovery of a new brain operation.

Reviewer #2 (Public review):

Summary:

This paper by Misra and Pessoa uses switching linear dynamical systems (SLDS) to investigate the neural network dynamics underlying threat processing at varying levels of proximity. Using an existing dataset from a threat-of-shock paradigm in which threat proximity is manipulated in a continuous fashion, the authors first show that they can identify states that each has their own linear dynamical system and are consistently associated with distinct phases of the threat-of-shock task (e.g., "peri-shock", "not near", etc). They then show how activity maps associated with these states are in agreement with existing literature on neural mechanisms of threat processing, and how activity in underlying brain regions alters around state transitions. The central novelty of the paper lies in its analyses of how intrinsic and extrinsic factors contribute to within-state trajectories and between-state transitions. A final set of analyses shows how the findings generalize to another (related) threat paradigm.

Strengths:

The analyses for this study are conducted at a very high level of mathematical and theoretical sophistication. The paper is very well written and effectively communicates complex concepts from dynamical systems. I am enthusiastic about this paper, but I think the authors have not yet exploited the full potential of their analyses in making this work meaningful toward increasing our neuroscientific understanding of threat processing, as explained below.

Weaknesses:

(1) I appreciate the sophistication of the analyses applied and/or developed by the authors. These methods have many potential use cases for investigating the network dynamics underlying various cognitive and affective processes. However, I am somewhat disappointed by the level of inferences made by the authors based on these analyses at the level of systems neuroscience. As an illustration consider the following citations from the abstract: "The results revealed that threat processing benefits from being viewed in terms of dynamic multivariate patterns whose trajectories are a combination of intrinsic and extrinsic factors that jointly determine how the brain temporally evolves during dynamic threat" and "We propose that viewing threat processing through the lens of dynamical systems offers important avenues to uncover properties of the dynamics of threat that are not unveiled with standard experimental designs and analyses". I can agree to the claim that we may be able to better describe the intrinsic and extrinsic dynamics of threat processing using this method, but what is now the contribution that this makes toward understanding these processes?

(2) How sure can we be that it is possible to separate extrinsically and intrinsically driven dynamics?

Author response:

The following is the authors’ response to the original reviews

Reviewer 1 (Public Review):

Summary

The manuscript uses state-of-the-art analysis technology to document the spatio-temporal dynamics of brain activity during the processing of threats. The authors offer convincing evidence that complex spatio-temporal aspects of brain dynamics are essential to describe brain operations during threat processing.

Strengths

Rigorous complex analyses well suited to the data.

Weaknesses

Lack of a simple take-home message about discovery of a new brain operation.

We have addressed the concern under response to item 1 in Recommendations for the authors of Reviewer 2 below.

Reviewer 1 (Recommendations for the authors):

The paper presents sophisticated analyses of how the spatiotemporal activity of the brain processes threats. While the study is elegant and relevant to the threat processing literature, it could be improved by better clarification of novelty, scope, assumptions and implications. Suggestions are reported below.

(1) Introduction: It is difficult to understand what is unsatisfactory in the present literature and why we need this study. For example, lines 57-64 report what works well in the work of Anderson and Fincham but do not really describe what this approach lacks, either in failing to explain real data in conceptual terms.

We have edited the corresponding lines to better describe what such approaches generally lack:

Introduction; Lines 63-66: However, the mapping between brain signals and putative mental states (e.g., “encoding”) remained speculative. More generally, state-based modeling of fMRI data would benefit from evaluation in contexts where the experimental paradigm affords a clearer mapping between discovered states and experimental manipulation.

(2) Also, based on the introduction it is unclear if the focus is on understanding the processing of threat or in the methodological development of experimental design and analysis paradigms for more ecologically valid situations.

In our present work, we tried to focus on understanding dynamics of threat processing while also contributing to methodological development of analysis of dynamic/ecologically inspired experiments. To that end, we have added a new paragraph at the end of Introduction to clarify the principal focus of our work:

Introduction; Lines 111-118: Is the present contribution focused on threat processing or methodological developments for the analysis of more continuous/ecologically valid paradigms? Our answer is “both”. One goal was to contribute to the development of a framework that considers brain processing to be inherently dynamic and multivariate. In particular, our goal was to provide the formal basis for conceptualizing threat processing as a dynamic process (see (Fanselow and Lester, 1987)) subject to endogenous and exogenous contributions. At the same time, our study revealed how regions studied individually in the past (e.g., anterior insula, cingulate cortex) contribute to brain states with multi-region dynamics.

(3) The repeated statement, based on the Fiete paper, that most analyses or models of brain activity do not include an exogenous drive seems an overstatement. There is plenty of literature that not only includes exogenous drives but also studies and documents them in detail. There are many examples, but a prominent one is the study of auditory processing. Essentially all human brain areas related to hearing (not only the activity of individual areas but also their communication) are entrained by the exogenous drive of speech (e.g. J. Gross et al, PLoS Biology 11 e1001752, 2013).

We have altered the original phrasing, which now reads as:

Introduction; Lines 93-95: Importantly, we estimated both endogenous and exogenous components of the dynamics, whereas some past work has not modeled both contributions (see discussion in (Khona and Fiete, 2022)).

Discussion; Lines 454-455: Work on dynamics of neural circuits in systems neuroscience at times assumes that the target circuit is driven only by endogenous processes (Khona and Fiete, 2022).

(4) Attractor dynamics is used as a prominent descriptor of fMRI activity, yet the discussion of how this may emerge from the interaction between areas is limited. Is it related to the way attractors emerge from physical systems or neural networks (e.g. Hopfield?).

This is an important question that we believe will benefit from computational and mathematical modeling, but we consider it beyond the scope of the present paper.

(5) Fig 4 shows activity of 4 regions, not 2 s stated in lines 201-202. Correct?

Fig. 4 shows activity of two regions and also the average activity of regions belonging to two resting-state networks engaged during threat processing (discussed shortly after lines 201-202). To clarify the above concern, we have changed the following line:

Results; Lines 228-230: In Fig. 4, we probed the average signals from two resting-state networks engaged during threat-related processing, the salience network which is particularly engaged during higher threat, and the default network which is engaged during conditions of relative safety.

(6) It would be useful to state more clearly how Fig 7B, C differs from Fig 2A, B (my understanding it is that in the former they are isolating the stimulus-driven processes)

We have clarified this by adding the following line in the Results:

Results; Lines 290-292: Note that in Fig. 7B/C we evaluated exogenous contributions only for stimuli associated with each state/state transition reported in Fig. 2A/B (see also Methods).

Reviewer 2 (Public Review):

Summary

This paper by Misra and Pessoa uses switching linear dynamical systems (SLDS) to investigate the neural network dynamics underlying threat processing at varying levels of proximity. Using an existing dataset from a threat-of-shock paradigm in which threat proximity is manipulated in a continuous fashion, the authors first show that they can identify states that each has their own linear dynamical system and are consistently associated with distinct phases of the threat-of-shock task (e.g., “peri-shock”, “not near”, etc). They then show how activity maps associated with these states are in agreement with existing literature on neural mechanisms of threat processing, and how activity in underlying brain regions alters around state transitions. The central novelty of the paper lies in its analyses of how intrinsic and extrinsic factors contribute to within-state trajectories and betweenstate transitions. A final set of analyses shows how the findings generalize to another (related) threat paradigm.

Strengths

The analyses for this study are conducted at a very high level of mathematical and theoretical sophistication. The paper is very well written and effectively communicates complex concepts from dynamical systems. I am enthusiastic about this paper, but I think the authors have not yet exploited the full potential of their analyses in making this work meaningful toward increasing our neuroscientific understanding of threat processing, as explained below.

Weaknesses

(1) I appreciate the sophistication of the analyses applied and/or developed by the authors. These methods have many potential use cases for investigating the network dynamics underlying various cognitive and affective processes. However, I am somewhat disappointed by the level of inferences made by the authors based on these analyses at the level of systems neuroscience. As an illustration consider the following citations from the abstract: “The results revealed that threat processing benefits from being viewed in terms of dynamic multivariate patterns whose trajectories are a combination of intrinsic and extrinsic factors that jointly determine how the brain temporally evolves during dynamic threat” and “We propose that viewing threat processing through the lens of dynamical systems offers important avenues to uncover properties of the dynamics of threat that are not unveiled with standard experimental designs and analyses”. I can agree to the claim that we may be able to better describe the intrinsic and extrinsic dynamics of threat processing using this method, but what is now the contribution that this makes toward understanding these processes?

We have addressed the concern under response to item 1 in Recommendations for the authors below.

(2) How sure can we be that it is possible to separate extrinsically and intrinsically driven dynamics?

We have addressed the concern under response to item 2 in Recommendations for the authors below.

Reviewer 2 (Recommendations for the authors):

(1) To address the first point under weaknesses above: I would challenge the authors to make their results more biologically/neuroscientifically meaningful, in particular in the sections (in results and/or discussion) on how intrinsic and extrinsic factors contribute to within-state trajectories and between-state transitions, and make those explicit in both the abstract and the discussion (what exactly are the properties of the dynamics of threat that are uncovered?). The authors may also argue that the current approach lies the groundwork for such efforts, but does not currently provide such insights. If they would take this position, that should be made explicit throughout (which would make it more of a methodological paper).

The SLDS approach provides, we believe, a powerful framework to describe system-level dynamics (of threat processing in the the present case). A complementary type of information can be obtained by studying the contribution of individual components (brain regions) within the larger system (brain), an approach that helps connect our approach to studies that typically focus on the contributions of individual regions, and contributes to providing more neurobiological interpretability to the results. Accordingly, we developed a new measure of region importance that captured the extent to which individual brain regions contributed to driving system dynamics during a given state.

Abstract; Lines 22-25: Furthermore, we developed a measure of region importance that quantifies the contributions of an individual brain region to system dynamics, which complements the system-level characterization that is obtained with the state-space SLDS formalism.

Introduction; Lines 95-99: A considerable challenge in state-based modeling, including SLDS, is linking estimated states and dynamics to interpretable processes. Here, we developed a measure of region importance that provides a biologically meaningful way to bridge this gap, as it quantifies how individual brain regions contribute to steering state trajectories.

Results; Lines 302-321: Region importance and steering of dynamics: Based on time series data and input information, the SLDS approach identifies a set of states and their dynamics. While these states are determined in the latent space, they can be readily mapped back to the brain, allowing for the characterization of spatiotemporal properties across the entire brain. Since not all regions contribute equally to state properties, we propose that a region’s impact on state dynamics serves as a measure of its importance.

We illustrate the concept for STATE 5 (“near miss”) in Fig. 8 (see Fig. S17 for all states). Fig. 8A shows importance in the top row and activity below as a function of time from state entry.The dynamics of importance and activity can be further visualized (Fig. 8B), where some regions of particularly high importance are illustrated together with the ventromedial PFC, a region that is typically not engaged during high-threat conditions. Notably, the importance of the dorsal anterior insula increased quickly in the first time points, and later decreased. In contrast, the importance of the periaqueductal gray was relatively high from the beginning of the state and decreased moderately later.

Fig. 8C depicts the correlation between these measures as a function of time. For all but STATE 1, the correlation increased over time. Interestingly, for STATES 4-5, the correlation was low at the first and second time points of the state (and for STATE 2 at the first time point), and for STATE 3 the measures were actually anticorrelated; both cases indicate a dissociation between activity and importance. In summary, our results illustrate that univariate region activity can differ from multivariate importance, providing a fruitful path to understand how individual brain regions contribute to collective dynamic properties.

Discussion; Lines 466-487: In the Introduction, we motivated our study in terms of determining multivariate and distributed patterns of activity with shared dynamics. At one end of the spectrum, it is possible to conceptualize the whole brain as dynamically evolving during a state; at the other end, we could focus on just a few “key” regions, or possibly a single one (at which point the description would be univariate). Here, we addressed this gap by studying the importance of regions to state dynamics: To what extent does a region steer the trajectory of the system? From a mathematical standpoint, our proposed measure is not merely a function of activity of a region but also of the coefficients of the dynamics matrix capturing its effect on across-region dynamics (Eichler, 2005; Smith et al., 2010).

How distributed should the dynamics of threat be considered? One answer to this question is to consider the distribution of importance values for all states. For STATE 1 (“post shock”), a few regions displayed the highest importance values for a few time points. However, for the other states the distribution of importance values tended to be more uniform at each time point. Thus, based on our proposed importance measure, we conclude that threat-related processing is profitably viewed as substantially distributed. Furthermore, we found that while activity and importance were relatively correlated, they could also diverge substantially. Together, we believe that the proposed importance measure provides a valuable tool for understanding the rich dynamics of threat processing. For example, we discovered that the dorsal anterior insula is important not only during high-anxiety states (such as STATE 5; “near miss”) but also, surprisingly, for a state that followed the aversive shock event (STATE 1; “post shock”). Additionally, we noted that posterior cingulate cortex, widely known to play a central role in the default mode network, to have the highest importance among all other regions in driving dynamics of low-anxiety states (such as STATE 3 and STATE 4; “not near”).

Methods; Lines 840-866: Region importance We performed a “lesion study”, where we quantified how brain regions contribute to state dynamics by eliminating (zeroing) model parameters corresponding to a given region, and observing the resulting changes in system dynamics. According to our approach, the most important regions are those that cause the greatest change in system dynamics when eliminated.

The SLDS model represents dynamics in a low dimensional latent space and model parameters are not readily available at the level of individual regions. Thus, the first step was to project the dynamics equation onto the brain data prior to computing importance values. Thus, the linear dynamics equation in the latent space (Eq. 2) was mapped to the original data space of N = 85 ROIs using the emissions model (Eq. 1):

where C† represents the Moore-Penrose pseudoinverse of C, and and denote the corresponding dynamics matrix, input matrix, and bias terms in the original data space.

Based on the above, we defined the importance of the ith ROI at time t based on quantifying the impact of “lesioning” the ith ROI, i.e., by setting the ith column of , the ith row of , and the ith element of to 0, denoted , ith ROI , and respectively. Formally, the importance of the was defined as:

where ‘∗’ indicates element-wise multiplication of a scalar with a vector, is the activity of ith ROI at time corresponds to the ith column of i is the inner product between th row of i and input th element of corresponds to the ith ROI. Note that the term is a function of both the i and th ROI’s activity as well as the coefficients of the dynamics matrix capturing the effect of region i on the one-step dynamics of the entire system (Eichler, 2005; Smith et al., 2010); the remaining terms capture the effect of the external inputs and the bias term on the one-step dynamics of the represents an indicator vector corresponding to the ith ROI.

After computing for a given run, the resultant importance time series was normalized to zero mean and unit variance.

(2) To address the second point under the weaknesses above: Given that the distinction between intrinsic and extrinsic dynamics appears central to the novelty of the paper, I would suggest the authors explicitly address this issue in the introduction and/or discussion sections.

The distinction between intrinsic and extrinsic dynamics is a modeling assumption of SLDS. We used such an assumption because in experimental designs with experimenter manipulated inputs one can profitably investigate both types of contribution to dynamics. While we should not reify the model’s assumption, we can gain confidence in our separation of extrinsically and intrinsically driven dynamics through controlled experiments where we can manipulate external inputs, or by demonstrating time-scale separation of intrinsic and extrinsic dynamics and that they operate at different frequencies. This is an important question that requires additional computational/mathematical modeling, but we consider it beyond the scope of the current paper. We have added the following lines in the discussion section:

Discussion; Lines 521-528: A further issue that we wish to discuss is related to the distinction between intrinsic and extrinsic dynamics, which is explicitly modeled in our SLDS approach (see Methods, equation 2). We believe this is a powerful approach because in experimental designs with experimenter manipulated inputs, one can profitably investigate both types of contribution to dynamics. However, complete separation between intrinsic and extrinsic dynamics is challenging to ascertain. More generally, one can gain confidence in their separation through controlled experiments where external inputs are manipulated, or by demonstrating timescale separation of intrinsic and extrinsic dynamics.

(3) In the abstract, the statement “.. studies in systems neuroscience that frequently assume that systems are decoupled from external inputs” sounds paradoxical after first introducing how threat processing is almost exclusively studied using blocked and event-related task designs (which obviously rely on external inputs only). Please clarify this.

In this work, we wished to state that the SLDS framework characterizes both endogenous and exogenous contributions to dynamics, whereas some past work has not modeled both contributions. To clarify, we have changed the corresponding line:

Abstract; Lines 19-20: Importantly, we characterized both endogenous and exogenous contributions to dynamics.

(4) In the abstract, the first mention of circles comes out of the blue; the paradigm needs to be introduced first to make this understandable.

We have rephrased the corresponding text:

Abstract; Lines 14-17: First, we demonstrated that the SLDS model learned the regularities of the experimental paradigm, such that states and state transitions estimated from fMRI time series data from 85 regions of interest reflected threat proximity and threat approach vs. retreat.

(5 In Figure 3, the legend shows z-scores representing BOLD changes associated with states. However, the z-scores are extremely low (ranging between -.4 and .4). Can this be correct, given that maps are thresholded at _p < .001 (i.e., z > 3._09)? A similar small range of z-scores is shown in the legend of Fig 5. Please check the z-score ranges.

The p-value threshold used in Fig. 3 is based on the voxelwise t-test conducted between the participantbased bootstrapped maps and null maps (see Methods : State spatial maps : “To identify statistically significant voxels, we performed a paired t-test between the participant-based boostrapped maps and the null maps.”). Thus, the p-value threshold in the figure does not correspond to the z-scores of the groupaveraged state-activation maps. Similarly in Fig. 5, we only visualized the state-wise attractors on a brain surface map without any thresholding. The purpose of using a z-score color bar was to provide a scale comparable to that of BOLD activity.

Version 2

Abstract

Abstract

In the past decades, functional MRI research has investigated task processing in a largely static fashion based on evoked responses during blocked and event-related designs. Despite some progress in naturalistic designs, our understanding of threat processing remains largely limited to those obtained with standard paradigms with limited dynamics. In the present paper, we applied Switching Linear Dynamical Systems to uncover the dynamics of threat processing during a continuous threat-of-shock paradigm. First, we demonstrated that the SLDS model learned the regularities of the experimental paradigm, such that states and state transitions estimated from fMRI time series data from 85 regions of interest reflected threat proximity and threat approach vs. retreat. After establishing that the model captured key properties of threat-related processing, we characterized the dynamics of the states and their transitions. Importantly, we characterized both endogenous and exogenous contributions to dynamics. The results revealed how threat processing can be viewed in terms of dynamic multivariate patterns whose trajectories are a combination of intrinsic and extrinsic factors that jointly determine how the brain temporally evolves during dynamic threat. Furthermore, we developed a measure of region importance that quantifies the contributions of an individual brain region to system dynamics, which complements the system-level characterization that is obtained with the state-space SLDS formalism. Finally, we investigated the generalizability of the modeling approach. The successful application of the SLDS model trained on one paradigm to a separate experiment illustrates the potential of this approach to capture fMRI dynamics that generalize across related but distinct threat-processing tasks. We propose that viewing threat processing through the lens of dynamical systems offers important avenues to uncover properties of the dynamics of threat that are not unveiled with standard experimental designs and analyses.

Introduction

Historically, systems neuroscience has sought to develop experimental paradigms where an effect of interest is varied while as many other variables remain fixed in an attempt to isolate the impact of the former while controlling for the latter. Such time-tested strategy has provided deep insights into brain and behavior. However, with the advent of neurotechnologies allowing the study of animals in more freely-behaving conditions, more naturalistic paradigms have been introduced (Hunt et al., 2021; Testard et al., 2024). At the same time, in human neuroimaging, researchers have defended the idea that dynamic and/or naturalistic paradigms are needed to advance our understanding of brain and behavior in a way that is not entirely constrained by rigid paradigms (Hasson and Honey, 2012; Sonkusare et al., 2019; Finn et al., 2020; Lee Masson and Isik, 2021; Grall and Finn, 2022). While such ideas resonate with those from ecological psychology and ethological approaches (Gothard et al., 2018; Mobbs et al., 2018; Evans et al., 2019), they introduce considerable challenges in the analysis of brain data and associated behaviors. If there are no specific trials or blocks in an experiment, what should constitute the unit of analysis? If the paradigm is more open-ended and lacks specific conditions (e.g., “attend to the visual stimulus” vs. “attend to the audio stimulus”), how should brain and behavior be investigated?

A promising approach is the use of unsupervised methods that consider time series dynamics (Anderson and Fincham, 2014; Vidaurre et al., 2016; Baldassano et al., 2017; Linderman et al., 2016; Taghia et al., 2017; Pandarinath et al., 2018; Townsend and Gong, 2018; Morioka et al., 2020; Wu et al., 2021; Singh et al., 2022; Song et al., 2023; Xu et al., 2023; Vahidi et al., 2024; Chen et al., 2024). In particular, Hidden Markov Models (HMMs) can be applied to time series data and partition it into “states” in time. For example, in one study, continuous narratives were presented and state transitions were identified from fMRI signals (Baldassano et al., 2017). Notably, state boundaries corresponded to human-annotated ones better than chance, and the match between model and human boundaries increased for more “high-level” brain regions (e.g., angular gyrus). Another study attempted to determine mental states while participants solved novel mathematical problems (Anderson and Fincham, 2014). Model states were interpreted as reflecting “encoding”, “planning”, “solving”, and “responding” stages based on fMRI data. Although the authors found evidence for the detected states, the study also highlights some of the challenges of applying unsupervised methods to complex paradigms. The estimated duration of the final responding state correlated well with the measured time participants spent entering their responses (r=0.53), providing some validation of that state. However, the mapping between brain signals and putative mental states (e.g., “encoding”) remained speculative. More generally, state-based modeling of fMRI data would benefit from evaluation in contexts where the experimental paradigm affords a clearer mapping between discovered states and experimental manipulation, as done here.

Overall, despite encouraging results (Anderson and Fincham, 2014; Vidaurre et al., 2016; Baldassano et al., 2017; Taghia et al., 2017, 2018; Meer et al., 2020; Wu et al., 2021; Song et al., 2021, 2023), the extent to which state-space models successfully parse fMRI data into meaningful mental/brain states remains poorly understood. Critically, while state-space models like HMM are able to segment time series data, they do not enable discovery of properties of state dynamics. In particular, how do brain signals evolve during a brain state that is maintained for a period of time? In systems neuroscience progress has been made in this context by studying brain signals via dynamical systems theory (Yu et al., 2005; Deco et al., 2008; Shenoy et al., 2013; Remington et al., 2018; Robson and Li, 2022). In particular, it has been suggested that attractor dynamics plays important computational functions (Miller, 2016; Khona and Fiete, 2022). However, comparable progress in fMRI research has lagged behind, perhaps because of the slowly evolving nature of the hemodynamic response (but see (John et al., 2022)).

Threat processing is typically studied in terms of discrete events and/or blocks that correspond to experimental conditions (but see (Mobbs et al., 2007, 2009; Murty et al., 2023; Levitas and James, 2024)). For example, a conditioned threat is presented and associated evoked brain responses determined (Fullana et al., 2016). But in many real-world situations, conditions involving threat unfold temporally in a less-discrete manner (Fanselow and Lester, 1987). In the present investigation, we investigated brain dynamics during a continuous threat-of-shock paradigm (Meyer et al., 2019; Limbachia et al., 2021) (Fig. 1). We sought to characterize the evolution of brain signals in terms of trajectories in multidimensional space, a framework at times called “computing with trajectories” (Buonomano and Maass, 2009). Instead of responses to discrete events possibly observed across multiple brain sites, we determined multivariate and distributed patterns of activity with shared dynamics.

Determining shared dynamics during threat processing.

Switching linear dynamical systems (SLDS) were used to model fMRI time series data (Yt) from a set of brain regions of interest (ROIs). The framework assumes that time series data can be segmented into a set of discrete states. The model represents brain signals in terms of a set of latent variables (Xt). For each state k the temporal evolution of the system is specified via a linear dynamical system with both intrinsic and input-related components. In the diagram, the system starts in state j (white star) and transitions to state k (the colored patches in the middle represent the subspaces associated with the two states). In state k, the system evolves according to the dynamics matrix Ak and input contributions (Vk). Overall, as states switch temporally, so does the corresponding linear dynamical system governing the system’s trajectory.

We employed a Switching Linear Dynamical Systems (SLDS) framework that estimates a generative model of the data with both states and transitions like HMMs (Ackerson and Fu, 1970; Barber, 2006; Fox et al., 2008; Linderman et al., 2019). In addition, each state is described in terms of a linear dynamical system that specifies the trajectory of brain activity during the state. Importantly, we estimated both endogenous and exogenous components of the dynamics, whereas some past work has not modeled both contributions (see discussion in (Khona and Fiete, 2022)). A considerable challenge in state-based modeling, including SLDS, is linking estimated states and dynamics to interpretable processes. Here, we developed a measure of region importance that provides a biologically meaningful way to bridge this gap, as it quantifies how individual brain regions contribute to steering state trajectories. Our approach demonstrated that we can recover key properties of threat-related processing in the brain (Grupe and Nitschke, 2013), while establishing novel properties of threat and safety dynamics. More generally, the approach can be applied to a broad range of continuous and/or naturalistic fMRI paradigms investigated in humans.

We also investigated model generalizability: how well does a model estimated in one paradigm generalize to related experiments? Examining this question is of importance because even statistically significant fits do not necessary generalize well to unseen data, not to mention to related but different paradigms. However, if an SLDS model captures meaningful spatiotemporal regularities associated with behavioral/brain states in a given dataset, the SLDS states should provide good fits to unseen data from a second, related experiment. Here, we investigated the extent to which the SLDS model that was fit to the moving circles paradigm generalized to a second, continuous paradigm involving threat and safety conditions.

Is the present contribution focused on threat processing or methodological developments for the analysis of more continuous/ecologically valid paradigms? Our answer is “both”. One goal was to contribute to the development of a framework that considers brain processing to be inherently dynamic and multivariate. In particular, our goal was to provide the formal basis for conceptualizing threat processing as a dynamic process (see (Fanselow and Lester, 1987)) subject to endogenous and exogenous contributions. At the same time, our study revealed how regions studied individually in the past (e.g., anterior insula, cingulate cortex) contribute to brain states with multiregion dynamics.

Results

Continuous threat processing switches between states

Participants watched two circles of different colors on the screen, at times moving close to each other, at times moving apart in a smooth but unpredictable fashion (Meyer et al., 2019; Limbachia et al., 2021). Upon circle collision, the participants received mild but unpleasant electrical stimulation together with an aversive sound. To enhance unpredictability and hence anxious apprehension by participants, the movement of the circles included multiple instances of “near misses” in which the circles nearly touched before retreating from each other.

Time series data were analyzed according to the SLDS framework (Fig. 1). Briefly, an SLDS estimates a model of the data using a set of unobserved (also called latent) variables assumed to govern system dynamics. The latent space is typically of considerably smaller dimension relative to the original data space (here 10 latent dimensions and 85 ROI time series, respectively). SLDSs consist of one linear dynamical system model per state, such that the trajectory of the system is governed by a single dynamical systems at at time. Like HMMs, an SLDS segments the data into a fixed number of states, such that the transitions between states follows a Markov process. At a given time, a single state is considered to be “active”. Overall, SLDS models provide a piecewise linear approximation of (potentially nonlinear) dynamics. Time series data were obtained from 85 regions involved in threat-related processing, as defined in a prior study (Murty et al., 2022).

The results reported below were obtained with 92 participants. A separate group of 30 participants was used to define hyperparameters, including the number of states (K = 6, see supplementary methods and Fig. S11) and the dimensionality of the latent space (D = 10, see supplementary methods and Fig. S10). Assessing statistical significance when complex models are applied to multi-participant data requires estimating variability of the estimates across participants. We employed a bootstrap procedure such that an SLDS model was estimated for each bootstrap sample, allowing us to estimate how SLDS parameters varied across the population (see Methods), thus allowing generalization beyond the current sample.

Model states are synchronous with the experimental paradigm

We evaluated if the states estimated by the model reflected experimental stimuli. For example, when the circles were far from each other, participants were “safe”, in contrast to when the circles were rather close and could progress to a collision. We encoded the input based on 20 bins defined in terms of circle proximity and direction: A1-A10 for approach, R1-R10 for retreat (see Methods, top inset in Fig. 2). When the circles approached each other, A1 indicated the farthest distance and A10 indicated that they collided; when they retreated from each other, R10 indicated that the circles collided and R1 indicated the farthest distance. We determined the associations between brain states and stimulus categories (i.e., bins) by calculating the probability of a brain state given the stimulus, P (state |stimulus) (see Methods). The probabilities are shown in Fig. 2A (e.g., P (STATE 2 | A10) = 0.63), where values exceeding those expected by chance are indicated in red outline. A total of five out of six brain states were significantly associated with specific stimuli categories. Examination of Fig. 2A can be used to provide a qualitative description of the estimated brain states. For example, STATE 2 was maximally observed when the circles were approaching and then collided, and thus reflects shock delivery (as well as peri-shock periods). As another example, STATE 5 was observed when the circles were in close proximity but did not touch. This state was particularly interesting given that the experiment was designed such that multiple “near miss” events would occur during the session, so as to enhance participants’ state of apprehension when the circles approached each other.

Brain states, state transitions, and input stimuli.

Stimuli were categorized into 20 bins based on circle distance and movement direction: A1-A10 for approach, R1-R10 for retreat (1 indicates circles are farthest and 10 indicates circle collision). (A) Table entries indicate the probability of being in a state given the input stimulus category. (B) Table entries indicate the probability of a state transition given the input stimulus category. In both tables, cells highlighted in red indicate states/transitions significantly associated with the corresponding stimulus category (p < 0.05, corrected for multiple comparisons). The color scale indicates probability.

We found that brain states were observed under specific stimulus conditions. But as the stimulus conditions change, do states transition in a systematic fashion? To evaluate this question, we determined the probability of a state transition given a specific stimulus category (Fig. 2B, e.g., P ([STATE 2 ↦ STATE 1] | A10) = 0.88; see Methods). Fig. 2B shows state transitions that were significantly associated with a stimulus category. For example, STATE 2 (centered on the collision event) transitioned to STATE 1 (“post-shock”) with very high probability for both stimulus categories A10 and R10 (i.e., when the circles collided). In Fig. 2B, to facilitate understanding, we ordered the rows by starting with the STATE 1 to STATE 3 transition (from “post-shock” to “not close”) and then chained the transitions in a manner that reflected the movement of the circles. Thus, the second row shows STATE 3 ↦ STATE 5 (from “not close” to “near miss”), and so on. For ease of reference, we summarize state descriptions in Box 1.

STATE 1: “post-shock”. Observed right after the circles collided; only state that followed STATE 2 in Fig. 2B.

STATE 2: “shock/peri-shock”. Observed during shock and peri-shock inputs.

STATE 3: “not near”. Observed when the circles were not near.

STATE 4: “not near”. Observed when the circles were not near and similar to STATE 3 in terms of inputs; STATE 4 followed STATE 5 when, after a near-miss collision, the circles continued to move apart from each other.

STATE 5: “near miss”. Observed when circles approached, came close to colliding, and then retreated.

Together, brain states and state transitions were systematically related to external stimuli, demonstrating that the model captured important properties of the experimental paradigm in an unsupervised manner.

Model states capture threat processing

We reasoned that if states detected by the model reflect known mental states, they should be associated with brain activity with coherent spatial patterns consistent with those observed with standard fMRI designs (such as block and event-related designs). SLDS states and state transitions are defined in a latent space of dimensionality considerably lower than the original brain data space (here 85 brain regions). Thus, to determine brain maps associated with specific states, we performed linear regression with a regressor that was on during the state in question and off otherwise (see Methods). Fig. 3 illustrates some of the states (see Fig. S13 for all states). For STATE 5, consistent with the “near miss” interpretation, mutiple brain areas linked to increased anxious apprehension were positively engaged, including the anterior insula and cingulate cortex (Mobbs et al., 2009, 2010; Alvarez et al., 2011; Grupe and Nitschke, 2013; Somerville et al., 2013; Meyer et al., 2019; Hur et al., 2020; Murty et al., 2022, 2023). STATE 4, which tended to follow STATE 5, exhibited a deactivation of these regions, consistent with the notion that participants experienced relative safety following a near collision (Schiller et al., 2008; Grupe and Nitschke, 2013; Somerville et al., 2013).

Voxelwise state activity maps and state contrast maps.

State activity maps for STATE 4 and STATE 5 and the contrast of the two states. Maps were corrected for multiple comparisons by thresholding voxels at p < 0.001 and at the cluster level at p < 0.05.

We further investigated the spatial properties of brain states by contrasting states. For example, the contrast of STATE 5 vs. STATE 4 revealed a pattern of activation very similar to that obtained in the literature for the contrast of higher vs. lower levels of threat. In particular, the contrast identified sectors of ventromedial PFC and posterior cingulate that have been linked to relative safety (Schiller et al., 2008; Grupe and Nitschke, 2013).

Temporal evolution of activity

In the threat paradigm, as the circles moved continuously, we expected that brain signals would also vary temporally. If SLDS states captured meaningful information, state transitions should be associated with relevant activity change. To evaluate this possibility, we determined signal evolution during temporal windows centered around state transitions (see Methods). Fig. 4 shows signal evolution in the same order as the state transitions of Fig. 2B in two key regions involved in threat-related processing, the dorsal anterior insula and the ventromedial PFC. For example, the transition from STATE 2 to STATE 1 reflected transitioning from a shock event to post-shock period. Accordingly, activity in the dorsal anterior insula, which is very sensitive to threat level (Mobbs et al., 2009, 2010; Alvarez et al., 2011; Grupe and Nitschke, 2013; Somerville et al., 2013; Meyer et al., 2019; Hur et al., 2020; Murty et al., 2022, 2023), underwent a drastic decrease. Likewise, when STATE 5 transitioned to STATE 4, activity in the dorsal anterior insula decreased, consistent with a transition from higher to lower threat and findings about responses in these two regions.

Region and network responses during state transitions.

Vertical lines indicate the time of state transition. State transitions are arranged in a chained manner: STATE 1 ↦ STATE 3 ↦ STATE 5 ↦ STATE 4 ↦ STATE 3 ↦ STATE 2 ↦ STATE 1 ↦ STATE 4 ↦ STATE 5 to facilitate continuity in reading responses across state transitions. Error bars correspond to the 95% confidence interval based on the standard error of the mean across participants.

In contrast, the transition from STATE 3 (“not close”) to STATE 2 (peri-shock/shock) was associated with a sharp increase in activity strength in the dorsal anterior insula.

SLDS models states and state transitions in a latent space, which makes it possible to investigate activity dynamics in regions not used for estimating the models, because once a sequence of states is determined one can examine activity in such regions at those times. Furthermore, activity can be determined at different spatial units, including ROI, voxel, or voxel cluster. In Fig. 4, we probed the average signals from two resting-state networks engaged during threat-related processing, the salience network which is particularly engaged during higher threat, and the default network which is engaged during conditions of relative safety. As anticipated, signal evolution during the transition STATE 5 ↦ STATE 4 was similar to that observed for the dorsal anterior insula and the ventromedial PFC, although signal evolution for the default network deviated to some extent from that of the ventromedial PFC (note that the ROIs and networks discussed here spatially overlapped; they were not selected to be independent).

Overall, the model captured regularities of the experimental paradigm reflected in systematic changes in activity levels across the brain. For example, state transitions STATE 3 ↦ STATE 5, STATE 4 ↦ STATE 3, and STATE 4 ↦ STATE 5 were associated with approaching circles (increased threat) and increased responses in the dorsal anterior insula and salience network following the state transition. Similarly, the transition STATE 5 ↦ STATE 4 was associated with retreating circles (decreased threat) and increased responses in the ventromedial PFC and default network following the state transition. Furthermore, responses of the right anterior insula and salience network during STATE 2 (“shock/peri-shock”) ramped up to the highest activity level with the STATE 3 ↦ STATE 2 transition from low to very high threat followed by a sharp decay with the STATE 2 ↦ STATE 1 transition that reflected the end of shock delivery.

Threat dynamics

Each state is characterized by a state-specific trajectory in latent space. Such state dynamics are defined by an intrinsic component and an external-input component, which additively define the system’s evolution (see Fig. 1 and Methods). In this section, we characterized the dynamics of threat processing based on state trajectories.

Intrinsic state dynamics evolve towards fixed-point attractors

First, we examined intrinsic state dynamics, namely the component of the dynamics that is separate from external input contributions. Does intrinsic dynamics converge to an attractor or does it diverge away from a fixed point? It is possible to test the stability of dynamics by evaluating properties of the dynamics matrix (matrix A; see Methods). We found that the intrinsic dynamics of all states studied were attractors (see supplementary Fig. S14).

An attractor state corresponds to a point in latent space. Such vector can be visualized as a map, which we call an “attractor spatial map” by projecting it onto the original brain data space in terms of ROI activity on a brain surface (Fig. 5; see Methods). For example, STATE 5 corresponded to “near miss” events and an attractor map with increased responses in the anterior insula/cingulate cortex and decreased responses in the ventromedial PFC and precuneus. The attractor for STATE 4 (“not near”) exhibited an opposite organization consistent with the context of increased safety/lower threat.

Attractor maps.

State attractors projected onto the space of regions of interest (85 ROIs) visualized on a brain surface map. At each ROI, the color scale represents activity strength at the attrator’s fixed point. ROI boundaries are marked in black.

Given that the SLDS model learns a generative model of the fMRI data during the experimental paradigm, state trajectories are based on both endogenous and exogenous contributions. How do these trajectories evolve relative to the state attractor? To investigate this question, we determined the vector field defined by the state’s endogenous dynamics, where vectors represent the direction and magnitude of how a system evolves if started at specific points (Fig. 6). Starting at a given point, trajectories evolved for multiple time steps in the overall direction of the attractor but after some time transitioned to other states (the green part of the trajectory indicates the mean lifetime of each state). The plots describe how, in a given state, the effective trajectory was determined by a combination of the endogenous flow and perturbations provided by the input. For example, in STATE 2 (shock/peri-shock) the vector field exhibited less curvature than several other states but trajectories were quickly steered away from the attractor because of inputs.

Evolution of state-specific trajectories.

Trajectories were determined in the latent space and projected onto a two-dimensional vector field for illustration (coordinate axes are specific to each state). The average trajectory across participants, starting at the white star, is initially shown in green and switches to red when the majority of trajectories (across participants) switch to another state. Whereas the trajectory includes endogenous and exogenous contributions to temporal evolution, the vector field represents the endogenous contribution only. The gray arrows indicate the effect of the state’s endogenous dynamics matrix, showing the direction and magnitude of evolution after a single time step. The blue cross indicates the state’s fixed point attractor; star indicates the start of the trajectory. PC: principal component.

Exogenous contributions to state trajectories

Following the qualitative description above (Fig. 6), we next quantified the contributions of external inputs in steering system trajectories. For each state, we determined the state’s centroid, namely, the centroid coordintate of all locations visited by the system during a specific state. Such geometrical location provided a reference position to investigate the input contributions. We considered the state centroid as reference location because we sought to determine how the inputs steered trajectories relative to the “most typical” position of a state.

We considered two scenarios. First, when the system was in state i, the input could drive the trajectory towards state i’s centroid or away from it (Fig. 7A). The “effective contribution” was determined by considering both direction and magnitude components (see Methods). We investigated the stimulus categories defined previously (A1—A10 and R10—R1). Consider STATE 1 (post-shock), for example (Fig. 7B). When the stimulus was R10 or R9, those inputs tended to push the system in the direction of STATE 1’s centroid; when the stimulus was R8, the trajectory was pushed away from the centroid by the input but very weakly. In addition, for STATE 2 (shock/peri-shock), we found that stimuli A8 to R10 pushed the system toward the centroid, but stimulus R9 moved the trajectory away from it. Overall, Fig. 7B displays how inputs steered state trajectories in the direction of the centroid (green cells) or away from it (purple cells). Note that in Fig. 7B/C we evaluated exogenous contributions only for stimuli associated with each state/state transition reported in Fig. 2A/B (see also Methods).

Effect of external inputs on steering state trajectories and driving state transitions.

(A) Left: Trajectory at t − 1 in state i. Inputs (colored arrows) could perturb the trajectory (translucent paths of the same color) by steering the evolution towards the state’s centroid (green) or away from it (pink). When the input directed it away from the state’s centroid, the input could push the system to switch into state j (red). Right: The input effect was measured via the cosine of the angle ϕ between the input vector and the line joining Xt−1 and state centroids. (B and C) Input effects with rows and columns representing states/state-transitions and input categories, respectively. Only states with significant association with inputs and only significant state transitions are shown (see Fig. 2). Cells with significant effects are highlighted in blue (p < 0.05, corrected for multiple comparisons).

Next, we investigated how inputs contributed to state transitions by steering trajectories in the direction of a new state’s centroid (see Methods). Consider the STATE 2 ↦ STATE 1 transition (i.e., from shock to post-shock). Input perturbations R10 and R9 provided the strongest push of the trajectory in the direction of STATE 1. In addition, when the system was in STATE 3 (“not near”), multiple inputs tended to direct the system’s evolution in the direction of STATE 2 (“shock/perishock”). Finally, when the system was in STATE 4 (“not close”) and the circles were approaching, inputs pushed the system in the direction of STATE 5 (“near miss”).

Taken together, the results show how external perturbations contributed to the evolution of state trajectories and state transitions.

Region importance and steering of dynamics

Based on time series data and input information, the SLDS approach identifies a set of states and their dynamics. While these states are determined in the latent space, they can be readily mapped back to the brain, allowing for the characterization of spatiotemporal properties across the entire brain. Since not all regions contribute equally to state properties, we propose that a region’s impact on state dynamics serves as a measure of its importance.

We illustrate the concept for STATE 5 (“near miss”) in Fig. 8 (see Fig. S17 for all states). Fig. 8A shows importance in the top row and activity below as a function of time from state entry. The dynamics of importance and activity can be further visualized (Fig. 8B), where some regions of particularly high importance are illustrated together with the ventromedial PFC, a region that is typically not engaged during high-threat conditions. Notably, the importance of the dorsal anterior insula increased quickly in the first time points, and later decreased. In contrast, the importance of the periaqueductal gray was relatively high from the beginning of the state and decreased moderately later. Fig. 8C depicts the correlation between these measures as a function of time. For all but STATE 1, the correlation increased over time. Interestingly, for STATES 4-5, the correlation was low at the first and second time points of the state (and for STATE 2 at the first time point), and for STATE 3 the measures were actually anticorrelated; both cases indicate a dissociation between activity and importance. In summary, our results illustrate that univariate region activity can differ from multivariate importance, providing a fruitful path to understand how individual brain regions contribute to collective dynamic properties.

Brain region importance.

(A) Importance measure and average BOLD signal for STATE 5 from state onset (t = 0) until the state’s average lifetime. (B) Average importance measure time series (left) and average BOLD signal for all ROIs during the average lifetime of STATE 5. (C) Spearman’s rank correlation coefficient between average BOLD signal (across participants) and average importance measure (across participants) from state onset (t = 0) until the average lifetime of STATE 5. PFC: Prefrontal cortex.

States explain threat processing in another threat-related experiment

If SLDS models capture spatiotemporal regularities associated with behavioral/brain states, determining states in a paradigm (such as the present moving circles paradigm) that shares important elements with a different experiment should be reflected when fitting the data of the latter. In other words, how well does a model estimated in one paradigm generalize to related experiments?

To test model generalizability, we applied the SLDS model estimated on the moving circles paradigm (call it MOV) to a separate experiment and dataset (call it ESC for “escape”) that investigated threat and safety processing. In the ESC experiment, participants manually controlled a rabbit icon using a joystick in a threat-avoidance task. The participant’s goal was to reach a safety area on the opposite end of the screen while avoiding a pursuing devil icon moving at varying speeds. If caught, participants received a mild but unpleasant electrical stimulus. Once a gate was successfully crossed, the participant entered a safe area (no shock was possible) where they remained for a few seconds (see Fig. 9A). Overall, the paradigm induced anxious apprehension during the chase phase and relief during safety.

Testing generalizability of the SLDS model.

(A) Experimental design of the “escape” (ESC) paradigm (data from a separate group of participants). Each trial started with a chase period (pink), where the participant controlled the rabbit icon to move towards the safety area (to their right) while being actively pursued by a threat icon. If the player entered the safety area without getting caught by threat, the safety gate closed, starting a safety period (green). Brain states were analyzed during windows centered around the time of entering the safe region (indicated by the vertical dashed line). (B) Table entries indicate the probability of being in a state given the time step relative to the window. Time steps marked in red and green indicate chase and safety periods respectively. (C) Table entries indicate the probability of a state transition given the time step. In both tables, cells highlighted in red indicate states/state transitions significantly associated with the window’s time step (p < 0.05, corrected for multiple comparisons). The color scale indicates probability values. No hemodynamic shift was applied to the tables in B and C. (D) Average fMRI response in the left Bed nucleus of the Stria Terminalis (BST). Error bars correspond to the 95% confidence interval based on the standard error of the mean across participants.

Model parameters inferred from the MOV dataset were used to decode states in the ESC dataset. We evaluated the extent to which MOV states could explain fMRI time series data of the esc design. To do so, we focused on a ten time-step long temporal window anchored around the event of entering safety (from −6.25 seconds to 5.0 seconds, with t = 0 indicating the time of entering safety; see Methods). We determined the probability of being in a given MOV state given the time step (Fig. 9B). The results indicated that STATE 5 (“near miss” in MOV) was expressed during the chase period; STATE 3 (“not near” in MOV) was detected from t = 0 seconds to the end of the temporal window. We also evaluated whether the system switched from STATE 5 to STATE 3 around t = 0 seconds by determining the probability of observing the transition STATE 5 ↦ STATE 3 as a function of time point (Fig. 9C; see Methods).

In the MOV paradigm, STATE 5 was associated with a state of apprehension likely experienced by participants during near-miss events. This is consistent with the finding that STATE 5 was detected during the chase phase of the ESC paradigm. STATE 3 (“not near” in MOV) occurred during the safety period (in ESC) and the transition from STATE 5 ↦ STATE 3 at t = 0. These findings indicate that the model trained on MOV was able to generalize in meaningful ways to the new paradigm.

To facilitate interpretation, no hemodynamic shifts were used in displaying the results shown in Fig. 9. Our choice was motivated by the fact that the ESC paradigm contained a discrete event at t = 0, namely, the transition into safety. However, one concern is that STATE 3 was already detected at safety onset, whereas safety-related processing triggered at t = 0 should be expressed in fMRI responses with some lag given that any responses triggered at t = 0 should translate to fMRI responses with a lag of around 2 seconds and peak around 4-6 seconds post event (Boyn- ton et al., 1996). Further analysis, however, helped explain the observed results, as outlined next. Average fMRI responses during the analyzed time window revealed a consistent response pattern across multiple brain regions sensitive to sustained threat in prior studies in the literature (Mobbs et al., 2009; Davis et al., 2010; Murty et al., 2023), as illustrated for the region of interest of the Bed nucleus of the Stria Terminalis (BST; Fig. 9D), a very important brain region involved in sustained threat processing (Davis et al., 2010). As shown, the BST response started decreasing before the end of the chase period, not at the end, suggesting that apprehension levels during the chase period decreased before the end of the chase period. To gain further insight concerning the response pattern of sustained-threat processing regions, we investigated the movement pattern of the chasing devil, which revealed a potential problem with the original experimental design. We programmed the chasing speed of the devil icon to be variable, but consistent across trials. In doing so, there was a consistent icon deacceleration around the halfway point of the chase, which unintentionally made the final phase of the chase less apprehension-inducing. In all, we believe the expression of STATE 3 (“not near” in MOV) right at entry into the safety area, as well as the transition from STATE 5 ↦ STATE 3, are actually consistent with a mental state of relative safety, as also reflected in the activation pattern in the BST (and other brain regions).

Discussion

In the present paper, we applied Switching Linear Dynamical Systems (SLDS) to uncover the dynamics of threat processing during a continuous threat-of-shock paradigm. First, we demonstrated that the SLDS model learned the regularities of the experimental paradigm, such that states and state transitions estimated from fMRI time series data from 85 ROIs reflected both the proximity of the circles and their direction (approach vs. retreat). After establishing that the model captured key properties of threat-related processing, we characterized the dynamics of the states and their transitions. The results revealed that the understanding of threat processing is enhanced by viewing it in terms of dynamic multivariate patterns whose trajectories are a combination of intrinsic and extrinsic factors that jointly determine how the brain temporally evolves during dynamic threat.

Past work has employed state-space models such as HMMs to estimate states from brain signals in an unsupervised fashion. Previous fMRI studies detected states with blocked design (Vi- daurre et al., 2018; Yamashita et al., 2021), where brain states are expected to align closely to the experimental structure. In continuous experimental paradigms, establishing brain states that meaningfully track experimental variables is considerably more challenging. In fact, in some studies employing HMMs with more unconstrained or naturalistic experiments, states were not explicitly mapped to stimulus events (Meer et al., 2020; Song et al., 2021, 2023). A recent interesting study applied HMMs to cell recordings in the rat PFC and found states that correlated with hide-and-seek behaviors, as well as states corresponding to previously undetected behaviors (Bagi et al., 2022). In general, although HMM models applied to brain data have shown to be a promising strategy to uncover unlabeled states, further work is needed to assess their performance in continuous paradigms in a manner that rigorously establishes the correspondence between states and external stimuli and/or behavior.

Critically, standard HMMs provide limited information about state dynamics, which is essential in understanding temporally unfolding brain processes supporting complex mental states and behaviors, such as the varying levels of threat and safety studied here. To address this gap, we applied the SLDS framework to our continuous threat-of-shock paradigm. There have been a few applications of SLDS models to brain data (Petreska et al., 2011; Linderman et al., 2019; Glaser et al., 2020), but as far as we know only one fMRI study investigating a blocked working memory task (Taghia et al., 2018). As one of the challenges with unsupervised state-space methods is that there is no ground truth concerning the states, here we first established that the model estimated states and transitions that captured the structure of the continuous paradigm, namely, circle proximity, shock events, and post-shock period. Of the six states detected by the model, five captured properties of the stimulus in a statistically significant manner. For example, the probability of being in STATE 2 given that the circles collided (input: A10) was 0.63, consistent with a state indicative of a “shock/peri-shock” period. State transitions were also associated with specific stimuli. More generally, the states captured the structure of the paradigm, which consisted of periods of increased and decreased threat given the proximity of the circles and the unpleasant shock delivered upon collision.

Further evidence that the states captured important features of threat-related processing was obtained by determining voxelwise brain maps for both individual states and state contrasts. For example, the contrast of STATE 5 (“near miss”) vs. STATE 4 (“not close”) revealed a pattern of activation very similar to that obtained in the literature for the contrast of higher vs. lower levels of threat (Mobbs et al., 2009, 2010; Alvarez et al., 2011; Grupe and Nitschke, 2013; Somerville et al., 2013; Meyer et al., 2019; Hur et al., 2020; Murty et al., 2022, 2023). In particular, the contrast identified sectors of ventromedial PFC and posterior cingulate that have been linked to relative safety (Schiller et al., 2008; Somerville et al., 2013; Grupe and Nitschke, 2013). In addition, state transitions reflected notable changes in activity levels, such as the transition of the lower-threat STATE 4 to the higher-threat STATE 5. Taken together, our analyses demonstrated that the model captured key properties of the continuous paradigm, building on previous studies using HMMs (Vidaurre et al., 2018; Meer et al., 2020; Song et al., 2023; Baldassano et al., 2017; Williams et al., 2022).

Our central goal was to discover and characterize the dynamics of continuous threat-related processing. From the perspective developed here brain regions jointly participate in state-specific trajectories. An important property of the SLDS framework is that it estimates state dynamics based on separate and additive endogenous and exogenous contributions. Determining the intrinsic dynamics is particularly informative because it uncovers how a brain state would evolve in the absence of additional inputs. In particular, does the state trajectory tend toward a fixed attractor, or does it evolve in some other more complex manner? Based on the properties of the endogenous state dynamics matrix, we determined that indeed all states behaved as attractors. But what do these attractors represent in terms of brain activity? To visualize the corresponding patterns, we determined “attractor maps” by projecting the model attractor (determined in the latent space) onto the brain space (times series data from ROIs). The attractor maps captured essential properties of the states determined by the model. For example, STATE 1 and STATE 2 maps exhibited inverse activity patterns, consistent with the fact that they largely reflected the post-shock and the shock/peri-shock periods, respectively. In addition, comparison of STATE 5 (“near miss”) and STATE 4 (“not near”) shows how distributed patterns with shared dynamics are involved in mental states of higher and lower anxious apprehension.

The endogenous dynamics can be further understood as providing a flow field that channels the temporal evolution of the system in the absence of inputs. However, in our continuous threat-of-shock paradigm, state trajectories were also determined by the inputs acting at each time step. Thus, visualizing the trajectory in terms of vector fields uncovered both contributions to system behavior. In combination with the information obtained from attractor maps, these results provide novel ways to understand the roles of key brain regions involved in threat-related processing. In addition to the qualitative information provided by vector field trajectories, we quantified the input contributions to states and state transitions. We determined how inputs contributed to trajectories in a state’s subspace and to transitions between states. For example, when the system was in STATE 2 (shock/peri shock), the input perturbation R10 steered the system in the direction of STATE 1 most vigorously, showing that soon after the circles collided, the brain tended to readily transition to a state of decreased activity in the insula and medial PFC. Overall, the analysis revealed how inputs steered a state’s trajectory toward or away from the state’s centroid, as well as how inputs directed a trajectory’s evolution in the direction of a new state.

Work on dynamics of neural circuits in systems neuroscience at times assumes that the target circuit is driven only by endogenous processes (Khona and Fiete, 2022). In other words, the external inputs do not vary significantly over time and therefore do not perturb state trajectories. While such a scenario may be adequate in some experimental paradigms, the assumption is problematic in cases where inputs vary continuously. In the present study, we did not assume that the brain was isolated from the environment. Instead, by parsing the dynamics in terms of endogenous and exogenous contributions, we were able to not only study intrinsic state attractors but also to characterize external inputs as perturbations that drove trajectories along meaningful directions, such as pulling or pushing inputs toward or away from state centroids. Additionally, we also identified input stimuli that effectively steered evolution in the direction of other state subspaces. We propose that our approach can be profitably employed in studying neuronal circuit dynamics in systems neuroscience more broadly.

In the Introduction, we motivated our study in terms of determining multivariate and distributed patterns of activity with shared dynamics. At one end of the spectrum, it is possible to conceptualize the whole brain as dynamically evolving during a state; at the other end, we could focus on just a few “key” regions, or possibly a single one (at which point the description would be univariate).

Here, we addressed this gap by studying the importance of regions to state dynamics: To what extent does a region steer the trajectory of the system? From a mathematical standpoint, our proposed measure is not merely a function of activity of a region but also of the coefficients of the dynamics matrix capturing its effect on across-region dynamics (Eichler, 2005; Smith et al., 2010).

How distributed should the dynamics of threat be considered? One answer to this question is to consider the distribution of importance values for all states. For STATE 1 (“post shock”), a few regions displayed the highest importance values for a few time points. However, for the other states the distribution of importance values tended to be more uniform at each time point. Thus, based on our proposed importance measure, we conclude that threat-related processing is profitably viewed as substantially distributed. Furthermore, we found that while activity and importance were relatively correlated, they could also diverge substantially. Together, we believe that the proposed importance measure provides a valuable tool for understanding the rich dynamics of threat processing. For example, we discovered that the dorsal anterior insula is important not only during high-anxiety states (such as STATE 5; “near miss”) but also, surprisingly, for a state that followed the aversive shock event (STATE 1; “post shock”). Additionally, we noted that posterior cingulate cortex, widely known to play a central role in the default mode network, to have the highest importance among all other regions in driving dynamics of low-anxiety states (such as STATE 3 and STATE 4; “not near”).

We believe the generalizability results presented in this study offer a valuable contribution to the ongoing discussion about the applicability of naturalistic paradigms in neuroscience. As highlighted in the Introduction, there is a growing interest in using more dynamic and ecologically valid experimental designs to advance our understanding of brain function. The successful application of the SLDS model, trained on the moving circles paradigm, to the separate escape experiment illustrates the potential of this approach to capture neural dynamics that generalize across related but distinct threat-processing tasks. This finding addresses a key challenge in naturalistic neuroscience: establishing that models derived from one paradigm can meaningfully explain brain activity in another. By showing that states identified in one threat-related task could be meaningfully mapped onto brain activity during a different threat-avoidance scenario, this study provides initial evidence that the SLDS framework can extract robust, task-general representations of threat processing, rather than idiosyncratic features of a particular experimental paradigm. Such crossparadigm validity (see also (Song et al., 2023)) will be crucial for building a more comprehensive and ecologically relevant understanding of how the brain processes and responds to dynamic threats in real-world scenarios.

Here, we also discuss some challenges and potential shortcomings of our study. The application of state-space models, including SLDS, to hemodynamic data is challenging because of the low-pass nature of the response. Specifically, the response to a brief event builds up to a peak around 4-6 seconds post onset and decays slowly over the next 4-8 seconds (Boynton et al., 1996). In the current study, we assumed a constant hemodynamic lag of approximately 5 seconds. Although our paradigm involved continuous movement of the circles, both shock events and transitions from approach to retreat (and vice versa) were discrete in nature. These events call for careful interpretation of the results given the potential for blending of hemodynamic responses. However, our experiment was designed such that most periods of circle approach and retreat lasted for 6-9 seconds so as to minimize consecutive approach/retreat transitions (Limbachia et al., 2021). Thus, the concern about the slow nature of the hemodynamic response was somewhat alleviated in the present context. In general, future studies are needed to develop SLDS approaches that can explicitly consider the properties of the hemodynamic response. Another limitation of the framework developed here is that we assumed, as in all state-model applications, that the underlying systems is well approximated by a series of states. With SLDS models this is less severe an assumption because one can consider the (possibly nonlinear) behavior of a system as modeled by a series of linear dynamical systems that approximate the overall behavior. Nevertheless, it will be valuable to investigate other approaches that address this concern, including considering that multiple states can coexist at a time (Greene et al., 2023). A further issue that we wish to discuss is related to the distinction between intrinsic and extrinsic dynamics, which is explicitly modeled in our SLDS approach (see Methods, equation 2). We believe this is a powerful approach because in experimental designs with experimenter manipulated inputs, one can profitably investigate both types of contribution to dynamics. However, complete separation between intrinsic and extrinsic dynamics is challenging to ascertain. More generally, one can gain confidence in their separation through controlled experiments where external inputs are manipulated, or by demonstrating time-scale separation of intrinsic and extrinsic dynamics.

In the present paper, we applied Switching Linear Dynamical Systems to uncover the dynamics of threat processing during a continuous threat-of-shock paradigm. First, we demonstrated that the model learned the regularities of the experimental paradigm, such that states and state transitions estimated from fMRI time series data from 85 ROIs reflected both the proximity of the circles and their direction (approach vs. retreat). After establishing that the model captured key properties of threat-related processing, we characterized the dynamics of the states and their transitions. The results revealed that threat processing can profitably be viewed in terms of dynamic multivariate patterns whose trajectories are a combination of intrinsic and extrinsic factors that jointly determine how the brain temporally evolves during dynamic threat. We propose that viewing threat processing through the lens of dynamical systems offers important avenues to uncover properties of the dynamics of threat that are not uncovered with standard experimental designs and analyses.

Methods

Moving Circles Dataset

Participants

One hundred and twenty-six participants (63 females, ages 18–30 years; average: 20.87, STD: 2.56) with normal or corrected-to-normal vision and no reported neurological or psychiatric disease were recruited from the University of Maryland community. The project was approved by the University of Maryland College Park Institutional Review Board and all participants provided written informed consent before participation. Data from four participants were not used due to technical issues. For the original publication of these data, see (Limbachia et al., 2021).

Out of the 122 participants, we used 30 to perform preliminary testing of the framework and defining model hyperparameters, including the number of states and the dimensionality of the latent space(see following subsections and supplementary). In order to avoid circularity in our analyses, data from this exploratory set were not used further. Thus, all results reported here are based on the held-out set of 92 participants.

Procedure and Stimuli

The experiment was designed to study the role of proximity on threat-of-shock processing. Participants watched two circles of different colors on the screen, at times moving close to each other, at times moving apart in a smooth but unpredictable fashion. Upon circle collision, the participants received a mild but unpleasant electrical stimulation together with an aversive sound. The movement of the circles was defined such that there were segments of approach and retreat that lasted between 2 and 9 seconds, with most segments lasting more then 6 seconds. To enhance unpredictability and hence anxious apprehension by participants, the movement of the circles included multiple instances of “near misses” in which the circles nearly touched (distance as close as 1.5 circle diameters between circle centers) before retreating away.

The experiment consisted of six 8-minute runs, each having two 3-minute blocks, during which the circles moved. The participants experienced on average 4 circle collisions and 7 near misses per run. All runs used different motion paths, but the visual stimulus was fixed across participants.

Finally, visual stimuli were presented using PsychoPy(http://www.psychopy.org/) and viewed on a projection screen via a mirror mounted to the scanner’s head coil. For further details about the experiment, the reader may refer to (Meyer et al., 2019; Limbachia et al., 2021).

Escape Dataset

Participants

A total of 57 subjects (24 females, ages 18-54; average: 20.5, STD: 1.9) were recruited from the University of Maryland, College Park, community. All subjects had normal or corrected-to-normal vision and reported no neurological disease or current use of psychoactive drugs. They provided written informed consent before participating in the study and were paid immediately after the experiment. The study was approved by the University of Maryland, College Park, Institutional Review Board.

Procedure and Stimuli

The experiment sought to study threat processing in presence of “imminent danger” followed by a period of safety. A total of eight 512 second-long runs were employed, each having 13.8±1.6 trials. Each trial began with a chase period that a followed by a safety period if the participant avoided being caught. Participants controlled a rabbit icon (“player”), which they could move on the screen at constant speed using a joystick. The player started at the left hand side of the screen and their objective was to reach a safety area located at the right end of the screen while avoiding a devil icon that pursued the player at time-varying speeds. Participants received an unpleasant but not painful electrical stimulation if the devil caught the player before the player entered the safety region. If the player successfully entered the safety area, this period lasted 6-10 seconds.

Regions of Interest

Time series data used for both datasets consisted of fMRI time series from 85 cortical and subcortical ROIs (defined as the average across voxels for each ROI). The ROIs were defined in a prior study by our group and intended to capture brain regions involved in threat-related processing (Murty et al., 2022).

Switching Linear Dynamical Systems

Switching Linear Dynamical Systems (SLDSs) are an example of a latent-variable framework that expresses the dynamics of observed signals using a set of unobserved variables assumed to govern system dynamics. The latent space is typically of considerable smaller dimension relative to the original data space (here 10 and 85, respectively). SLDSs consist of one linear dynamical system model per state, such that the trajectory of the system is governed by a single dynamical system at at each time step. Switching between states is assumed to follow a Markov process. Overall, SLDS models provide a piecewise linear approximation of (potentially nonlinear) dynamics.

Consider activity Yt ∈ ℝN, t = 1, 2, …, T (T time steps is the duration of a typical run) from N = 85 brain ROIs during an experiment described using M input variables Ut ∈ ℝM, t = 1, 2, …, T. The model represents brain signals Yt in a low dimensional subspace as a set of D latent variables Xt ∈ ℝD, t = 1, 2, …, T (D < N). Observed BOLD activity activity is related to the latent variables using a linear transformation, such that where C ∈ ℝN×D is the observation matrix, d ∈ ℝN is a bias term, Σv ∈ ℝN×N is a noise covariance matrix and 𝒩 (μ, Σ) is a multivariate Gaussian distribution with mean μ and covariance Σ. The expression above is called the observation step.

Latent trajectory dynamics is modeled using a set of K linear dynamical systems that are indexed by a time varying switching variable Zt ∈ {1, 2, …, K}, which we call the state. At each time step t only the linear dynamical system corresponding to the state Zt drives the temporal evolution of system, such that where is the linear dynamics matrix, is the input matrix reflecting the experimental paradigm variables, is a bias term and is a noise covariance matrix. The expression above is called the dynamics step.

State transition follows a modification of the Markov property, where the probability of switching to state Zt from the previous state Zt−1 depends on the previous values of the state Zt−1 and the latent variables at t − 1, Xt−1: where rj ∈ ℝD and Q ∈ RK×K is the Markovian transition matrix, whose ith row and jth column element indicates the transition probability from state i to state j. This expression above is called the state switching step. rj is a weight vector (indexed by the state label, j) that specifies the additional dependency of Zt on Xt−1 (on top of the Markov dependence of Zt−1 on Zt), allowing state transitions based on the location of the trajectory in the latent space. SLDS with the above additional dependency is also known as recurrent SLDS (Linderman et al., 2016) and is shown to be more suitable to explain realistic timeseries than SLDS without the recurrent dependencies (Linderman et al., 2019). Note that, on removing rj from the above equation the state transition probability is the same as that of a Markov model.

A group-level SLDS model was estimated based on the data of 92 participants that participated in the moving circles experiment by using a variational Laplace Expectation Maximization algorithm (Linderman et al., 2016). Specifically, all of the following were estimated: . In this manner, we obtained model variables Xt and Zt. We employed the SLDS implementation provided in the ssm python library (https://github.com/lindermanlab/ssm).

Modeling experimental stimuli as SLDS inputs

To encode the experimental paradigm, we categorized the stimuli based on the proximity and direction of motion of the moving circles. The instantaneous proximity was calculated as the normalized Euclidean distance between the circles, ranging from 0 (farthest) to 1 (touching). Direction was calculated based on the discrete difference of proximity: +1 (approach) or −1 (retreat). Proximity values were quantized into 10 equal-sized bins ([0.0, 0.1), [0.1, 0.2), ⋯, [0.9, 1.0]), one per binning per direction, resulting in 20 stimulus categories: A1, A2, ⋯, A10 for approach; R1, R2, ⋯, R10 for retreat). The stimulus conditions were thus encoded as a 20-dimensional binary vector Ut at each time step t, with a 1 corresponding to the active category and 0 elsewhere. The number of stimulus categories was arbitrarily defined, but we verified that key results were robust to the number of input bins (see Supplementary Fig. S15 and S16).

For a state k, the SLDS model represents the contribution of input on dynamics using the term VkUt (Eq. 2), where Ut encodes the stimulus and Vk maps the input into the latent space. Note Vk is state-specific, allowing distinct representations of the same stimulus category per state.

Assessing statistical significance using bootstrapping

Data from the held-out set of 92 participants of the moving circles dataset) were used to report findings of the study and determine statistical significance, which was determined via bootstrapping (Efron and Tibshirani, 1994). A total of 500 bootstrap samples of the dataset were generated by sampling participants with replacement, and an SLDS model was fit on each bootstrap iteration. Because state labels are not necessarily preserved across bootstrap samples, we realigned them using a combination of the Hungarian algorithm (Kuhn, 1955) and k-means clustering. To do so, for each bootstrap sample, state parameters were concatenated to form K = 6 “large vectors” θ, one for each state. The resulting vectors were then clustered into K clusters. The K cluster centroids provided representative vectors θ for each state. In this manner, aligning state labels across bootstrap samples amounted to assigning each bootstrap sample’s θ vectors to the K cluster centroids. The assignment problem was determined by the Hungarian assignment algorithm, such that the bootstrap sample’s state θ vectors were maximally aligned to the K cluster centroids.

To determine the number of bootstrap iterations, we generated samples in steps of 100 until the estimates of the above mentioned θ centroids were stable. Specifically, the elements of each centroid did not change (within 10−5 precision) upon adding bootstrap samples. The procedure was stopped at 500 bootstrap samples.

Calculating the probability of a state given a stimulus

To determine P (state| input), we defined a state by input table. Each entry corresponded to the proportion of time a state Zt assumed a specific value z given that the input Ut assumed a specific value u. This counting procedure can be briefly summarize in the (somewhat awkward) expression where 𝟙(·) represents an indicator function that takes value 1 when the operand is true and 0 otherwise. Given the hemodynamic delay (approximately 2 secs to initiate a response and 4-6 secs to peak; (Boynton et al., 1996)), we used a time shift of 3 time steps (3.75 seconds) in considering the relationship between SLDS state and stimulus category. That is, to facilitate interpretation, we assumed that the effect of circle proximity (as encoded in terms of the discrete stimulus categories) would be expressed in the fMRI timeseries 3 time steps later.

We tested the statistical significance of each entry in the state by input table as follows. To account for sampling variability of the sample we studied, we computed a table for each iteration of the bootstrap, allowing us to estimate the variability of our results across the population (group level). Similarly, we obtained surrogate tables, one for each bootstrap resample, by randomly shifting the state sequences by random offsets. In this manner, the relative order of the states was randomized with respect to order of stimuli but the other properties were maintained (transition structure in the sequences). By obtaining bootstrap and surrogate tables; for each table cell corresponding to a given state and stimulus, we obtained a population-level distribution of table values and a null distribution.

Based on these results, each table entry mean was compared to the mean of the corresponding null distribution via a paired Wilcoxon signed rank test at p < 0.05. To correct for multiple comparisons across the number of states and stimulus categories (6 × 20), Bonferroni correction was applied.

Calculating the probability of a state transition given a stimulus

State transitions were linked to the stimuli using an analogous procedure as that to map states to stimuli. We created a table with 8 state transitions of interest (see supplementary methods, Fig. S12) by 20 stimulus categories. Each table entry was calculated by counting the number of times a given state transition was observed given a stimulus category. Formally, this counting procedure can be succinctly as follows: for a row r representing transition state i ↦ state j and column u, the corresponding table entry measures the probability where 𝟙(·) represents an indicator function. Once again, this is calculated by considering the hemodynamic delay. Finally, statistical inference was performed using a similar approach as described previously for calculating probabilities of states given stimuli.

State spatial maps

We sought to characterize the spatial organization of brain states: for a given state k, what is the typical pattern of activity throughout the brain. This could be accomplished by using multiple regression with variables for each state: 1 when the state is on, 0 otherwise. We performed multiple regression in this manner in a voxel-wise manner. Note that this was possible because once the states are estimated (based on 85 ROIs), the regression can be applied at any spatial level desired (ROI or voxel). The regression model can be expressed mathematically as where 𝟙(Zt = k) is an indicator function denoting whether state k is active at time t, βk is the corresponding regression coefficient, and εt is an error term. For notational simplicity, terms accounting for confounds like scanner drift and head motion have been omitted from the above expression but were included in the regression model.

State maps were evaluated statistically based on the bootstrap approach discussed previously. First, state activity maps were determined for every participant, separately. A group-level map was then obtained by averaging across participants. This procedure was performed for every bootstrap sample, allowing us to estimate sampling variability. Similarly, a set of null state maps was generated by repeating the same procedure but with randomly permuted state sequences. To identify statistically significant voxels, we performed a paired t-test between the participant-based boot-strapped maps and the null maps. To account for multiple comparisons, we employed a cluster-based approach (Friston et al., 1994). Voxels were first thresholded at p < 0.001, and clusters with at least c = 18 voxels were deemed statistically significant at the cluster level. To determine the minimum cluster size c, we employed null state maps. The cluster size c was such that only 5% of the clusters in the null state maps would be detected, corresponding to a 5% false-positive rate.

For contrast maps, an analogous approach was employed, but at the individual level contrast maps were determined by computing βi − βj, where i and j are two states in question.

Responses during state transitions

Responses at a given ROI were evaluated during temporal windows centered around state transitions. For transitions from state i to state j, we chose windows where the first half of the time steps were labelled as state i and the final time steps as state j. We chose windows of length 8 time steps such that they were long enough to reveal activity changes but also observed frequently enough to generate a reliable estimate. For a given state transition, participant-level responses were determined by averaging responses across all qualifying time windows. To estimate variability, group-level responses were determined by averaging participant-level responses for each bootstrap sample.

We also determined responses at the “network” level, which were obtained by averaging ROI responses across all ROIs inside the network. We considered the salience (ventral attention) and default networks, as defined in the Schaefer 100-region 7-network parcellation (Schaefer et al., 2017).

State specific intrinsic dynamics and fixed points

Each state’s dynamics in the latent space was modelled using a linear dynamical system, which evolves as follows:

To characterize the intrinsic dynamics exclusively, assume that the input to the system is zero across time:

Assuming that the dynamics are stable and approach an equilibrium point as a function of time (also known as a “fixed point” or attractor), we obtain where I is a D × D identity matrix. All trajectories in the neighbourhood of the fixed point (also called the basin of attraction) will eventually converge towards the fixed point.

Testing for stable intrinsic dynamics

The dynamics of a linear dynamical system is determined by the eigenvalues of the dynamics matrix Ak. In particular, the dynamics is stable if all eigenvalues lie inside a unit circle on the complex plane (refer to (Luenberger, 1979; Robinson, 2012) for detailed accounts on stability of linear dynamical systems). Therefore it suffices to test whether norm of the largest eigenvalue is less than unity. Bootstrapping was again used to evaluate the statistical significance of the results. We determined largest eigenvalues for all bootstrap samples, and calculated the p-value as the fraction of the eigenvalues with absolute value greater than 1. States with p-value < 0.05 were considered to follow stable dynamics. Bonferroni correction was applied to correct for multiple comparisons across K states.

Visualizing state attractors on the brain

Given an attractor state i, its fixed point is a vector in the D-dimensional latent space. For the purpose of visualizing state attractors on the brain, we projected them onto the original data space (the time series space of 85 ROIs) using the observation step of the SLDS model (Eq. 1).

Where is state i’s attractor in ROI space. State attractors in ROI space were estimated across bootstrap samples, and the average pattern was visualized. The average attractor pattern was thresholded at each ROI (p < 0.05) based on a Wilcoxon signed-rank test (p < 0.05); that is, by comparing the activity levels to zero. Bonferroni correction was applied to correct for multiple comparisons across K states and 85 ROIs.

State trajectories and vector fields

For each state, trajectories were determined first in the latent space (D = 10). We averaged (across participants) all windows with 10 time steps starting at a specific state k. Note that trajectories correspond to Eq. 4 and consider both endogenous and exogenous terms. At each time step of the average trajectory, a trajectory could continue along a state or switch to a state j ≠ k. The state label for a time step along the trajectory was assigned to the most typical state at that time step; i.e., the modal state. We colored the trajectory green when the state label was k and red otherwise (state j ≠ k). To illustrate the trajectory, we plotted it in 2D. To do so, we fit principal component analysis on the 10 points constituting the average trajectory and projected it onto the plane spanned by the top two principal components.

Trajectories were plotted on vector fields that represent how a particle would flow under the endogenous dynamics (i.e., based on Ak). This allowed us to visualized how external inputs contributed to the evolution of the system (note that, by definition, without inputs trajectories would converged to the state’s attractor). At each point on the 2D plane, the vectors represent the trajectory evolution for a single time step. Vector field points were sampled uniformly on the 2D plane. The points were projected to the latent space using the top two principal components above, evolved according to endogenous dynamics rule (Eq. 5), and projected back onto the 2D plane again using the top two principal components. Finally, the state’s fixed point attractor (a point in the latent space) was also projected onto the 2D plane in the same manner.

Quantifying contribution of inputs in state dynamics

As described above, state k’s trajectory follows the following dynamical rule given in Eq. 4, with the term VkUt representing contribution of external inputs as a vector in latent space. To investigate the contribution of external inputs to the latent trajectory, we determined how inputs steered the system’s trajectory. For each state, we determined the state’s centroid, namely, the vector Xc,k that represented the centroid coordinate of all locations visited by the system during state k. Such geometrical location provided a reference position to investigate the input contributions.

Given a state, the external input at a given time can either drive latent trajectories towards the state centroid or away from it. To quantify input effects, we measured the length of the projection of the input vector along the direction connecting the position of the state trajectory and the state centroid: where ⟨·, ·⟩ indicates the inner product between two vectors and ‖·‖ indicates the Euclidean norm. Thus, positive (negative) leff values indicate that the input pushes state trajectories towards (away from) the state centroid. We focused on inputs associated with the given state as defined throughout the paper (methods for mapping states to stimuli, see Methods). Accordingly, the input effects summarised in a table where the (i, j)th element represents the effect of input j on trajectories in state i.

Tables were constructed as previously such that each bootstrap iteration generated one instance of the table. The results shown correspond to the average value across bootstrap samples. Statistically, we tested whether the average input effects were significantly different from zero by conducting a Wilcoxon’s signed-rank test at each table entry (p < 0.05). Bonferroni correction was applied to correct for multiple comparisons across K = 6 states and 20 input categories.

Quantifying contribution of inputs in state transitions

Next, we quantified the contributions of input stimuli to state transitions. Specifically, for particular state transitions i ↦ j, we determined the extent to which input stimuli contributed to such transitions. As above, leff was used to measure the effect of the external input vector along the direction joining the position of the state trajectory and state j’s centroid (in place of state i’s centroid). Positive values indicate that the input pushed the latent vector from state i towards state j.

Tables were constructed as previously, such that each bootstrap iteration generated one instance of the table. The results shown correspond to the average value across bootstrap samples. Statistically, we tested whether average input effects were significantly different from zero by conducting a Wilcoxon’s signed-rank test at each table entry (p < 0.05). Bonferroni correction was applied to correct for multiple comparisons across 8 state transitions of interest and 20 input categories.

Testing SLDS model generalizability

After fitting the SDLS model based on the data of the moving circles (MOV) paradigm, we investigated if the model states generalized to another threat-related paradigm, namely, the escape (ESC) experiment described above (see Fig 9A). Specifically, after fitting the data to the first dataset, we determined the variables Xt and Zt for the ESC dataset. ESC time series data was analyzed for temporal windows centered around the time step when the player entered safety (rounded to the nearest data sample). Windows straddled between chase and safety periods, from t = −6.25 to t = 5.0 with t = 0 being the cross-into-safety point.

To determine the best-matching MOV states for the ESC time series data, we quantified P (state | time step) for all combinations of state by time steps see Fig. 9. Each value was computed as the normalized frequency of observing a state Zt at a given time step twin of the window. Formally: where 𝟙(·) represents the indicator function.

We determined the statistical significance of the probability values as follows. To determine the probabilities that would be observed by chance, we determined surrogate tables (states by time step) by shifting the MOV state sequences by random offsets (analogous to the approach used in mapping states to stimuli in MOV, see Methods), which provided a null distribution of probability values. Each probability value was then compared to the mean of the corresponding null distribution via a paired Wilcoxon signed rank test at p < 0.05. To correct for multiple comparisons across the number of states and time steps (6 × 10), Bonferroni correction was applied. In a similar fashion, the probability of a state transition from STATE 5 ↦ STATE 3 was tested for statistical significance (analogous to mapping state transitions to stimuli in MOV, see Methods).

Region importance

We performed a “lesion study”, where we quantified how brain regions contribute to state dynamics by eliminating (zeroing) model parameters corresponding to a given region, and observing the resulting changes in system dynamics. According to our approach, the most important regions are those that cause the greatest change in system dynamics when eliminated.

The SLDS model represents dynamics in a low dimensional latent space and model parameters are not readily available at the level of individual regions. Thus, the first step was to project the dynamics equation onto the brain data prior to computing importance values. Thus, the linear dynamics equation in the latent space (Eq. 2) was mapped to the original data space of N = 85 ROIs using the emissions model (Eq. 1): where C† represents the Moore-Penrose pseudoinverse of C, and and denote the corresponding dynamics matrix, input matrix, and bias terms in the original data space.

Based on the above, we defined the importance of the ith ROI at time t based on quantifying the impact of “lesioning” the ith ROI, i.e., by setting the ith column of ,the ith row of ,and the ith element of to 0, denoted , and respectively. Formally, the importance of the ith ROI was defined as: where ‘∗’ indicates element-wise multiplication of a scalar with a vector, is the activity of ith ROI at time , corresponds to the ith column of is the inner product between ith row and input corresponds to the ith element of and 𝟙 ∈ ℝN represents an indicator vector corresponding to the ith ROI. Note that the term is a function of both the ith ROI’s activity as well as the coefficients of the dynamics matrix capturing the effect of region i on the one-step dynamics of the entire system (Eichler, 2005; Smith et al., 2010); the remaining terms capture the effect of the external inputs and the bias term on the one-step dynamics of the ith ROI.

After computing for a given run, the resultant importance time series was normalized to zero mean and unit variance.